---
output: github_document
bibliography: "inst/REFERENCES.bib"
csl: "inst/national-institute-of-health-research.csl"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "75%"
)
```

# rdecision

<!-- badges: start -->
<!-- badges: end -->

The goal of `rdecision` is to provide methods for assessing health care 
interventions using cohort models (decision trees). Mechanisms are provided for 
associating an uncertainty distribution with each source variable and for
ensuring transparency of the mathematical relationships between variables.
The package terminology follows Briggs *et al* "Decision Modelling for Health
Economic Evaluation" [@briggs2006].


## Installation

You can install the released version of rdecision from 
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("rdecision")
```

## Example

Consider the fictitious and idealized decision problem of choosing between
providing two 
forms of lifestyle advice, offered to people with vascular disease, which reduce
the risk of needing an interventional procedure. The model has a time horizon of 
1 year. The cost to a healthcare provider of the interventional procedure 
(e.g. inserting a stent) is 3000 GBP; the cost of providing the current form of
lifestyle advice, an appointment with a dietician ("diet"), is 50 GBP and the
cost of providing an alternative form, attendance at an exercise
programme ("exercise"), is 500 GBP. If the advice programme is successful, there
is no need for an interventional procedure. In a small trial of the "diet" 
programme, 12 out of 68 patients avoided having a procedure, and in a separate 
small trial of the "exercise" programme 18 out of 58 patients avoided the
procedure. It is assumed that the baseline characteristics in the two trials
were comparable, that the model is from the perspective of the healthcare
provider and that the utility is the same for all patients.

A decision tree can be constructed to estimate the uncertainty of the cost 
difference between the two types of advice programme, due to the finite sample 
sizes of each trial. The proportions of each advice programme being successful
(i.e. avoiding intervention) are represented by model variables with 
uncertainties which follow Beta distributions. Probabilities of the failure
of the programmes are calculated using expression model variables to ensure
that the total probability associated with each chance node is one.

```{r pvars, echo=TRUE}
library("rdecision")
p.diet <- BetaModVar$new("P(diet)", "", alpha=12, beta=68-12)
p.exercise <- BetaModVar$new("P(exercise)", "", alpha=18, beta=58-18)
q.diet <- ExprModVar$new("1-P(diet)", "", rlang::quo(1-p.diet))
q.exercise <- ExprModVar$new("1-P(exercise)", "", rlang::quo(1-p.exercise))
```

The decision tree is constructed from nodes and edges as follows:
```{r tree, echo=TRUE}
t.ds <- LeafNode$new("no stent")
t.df <- LeafNode$new("stent")
t.es <- LeafNode$new("no stent")
t.ef <- LeafNode$new("stent")
c.d <- ChanceNode$new("Outcome")
c.e <- ChanceNode$new("Outcome")
d <- DecisionNode$new("Programme")

e.d <- Action$new(d, c.d, cost = 50, label = "Diet")
e.e <- Action$new(d, c.e, cost = 500, label = "Exercise")
e.ds <- Reaction$new(c.d, t.ds, p = p.diet, cost = 0, label = "success")
e.df <- Reaction$new(c.d, t.df, p = q.diet, cost = 5000, label = "failure")
e.es <- Reaction$new(c.e, t.es, p = p.exercise, cost = 0, label = "success")
e.ef <- Reaction$new(c.e, t.ef, p = q.exercise, cost = 5000, label = "failure")

DT <- DecisionTree$new(
  V = list(d, c.d, c.e, t.ds, t.df, t.es, t.ef),
  E = list(e.d, e.e, e.ds, e.df, e.es, e.ef)
)
```

```{r treedraw,echo=FALSE,results="hide",fig.keep="last",fig.align="center"}
DT$draw(border = TRUE)
```

```{r pointestimate, echo=FALSE}
RES <- DT$evaluate()
```
The expected per-patient cost of each option is obtained by evaluating the
tree with expected values of all variables using `DT$evaluate()`. Examination
of the data frame from this call shows that the expected per-patient cost of
the diet advice programme is `r round(RES$Cost[RES$Programme=="Diet"],2)` GBP 
and the per-patient cost of the exercise programme 
is `r round(RES$Cost[RES$Programme=="Exercise"],2)` GBP, a point estimate saving
of
`r round(RES$Cost[RES$Programme=="Diet"]-RES$Cost[RES$Programme=="Exercise"],2)`
per patient if the exercise advice programme is adopted. 

```{r psa, echo=FALSE}
RES <- DT$evaluate(setvars="random", N=1000)
RES <- reshape(RES, idvar="Run", timevar="Programme", direction="wide")
RES$Difference <- RES$Cost.Diet - RES$Cost.Exercise
```

The confidence interval of the
cost saving is estimated by repeated evaluation of the tree, each time sampling
from the uncertainty distribution of the two probabilities using, for example, 
`DT$evaluate(setvars="random", N=1000)` and inspecting the resulting data frame.
From 1000 runs, the 95% confidence interval of the per patient cost saving 
is `r round(quantile(RES$Difference, probs=c(0.025)),2)` GBP 
to `r round(quantile(RES$Difference, probs=c(0.975)),2)` GBP and it can be
concluded that there is no evidence for there being a cost difference between
the choices.

# Acknowledgements
In addition to using base R [@rcoreteam2020], `redecision` relies heavily on
the `R6` implementation of classes [@chang2020] and the `rlang` package for
error handling and non-standard evaluation used in expression model variables
[@henry2020]. Building the package vignettes and documentation rely on the
`testthat` package [@wickham2011], the `devtools` package [@wickham2020] and
`rmarkdown` [@xie2018a,@allaire2020,@xie2020].

Underpinning graph theory is based on terminology, definitions and
algorithms from Gross *et al* [-@gross2013], the Wikipedia 
glossary [@wikipedia2021] and links therein. Topological sorting of graphs
is based on Kahn's algorithm [-@kahn1962]. Some of the terminology for decision
trees was based on the work of Kamiński *et al* [-@kaminski2018] and an 
efficient tree drawing algorithm was based on the work of Walker [-@walker1989]. 
Terminology for decision trees in health economic evaluation was based on
the book by Briggs *et al* [-@briggs2006] and the output format and terminology 
follows ISPOR recommendations [@briggs2012a]. 

Citations for examples used in vignettes are given in applicable vignette
files. 

# References

