---
output: github_document
bibliography: "inst/REFERENCES.bib"
csl: "inst/nature-no-et-al.csl"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r}
#| include = FALSE,
#| purl = FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  comment = "#>"
)
# read vignette build utility functions
knitr::read_chunk(file.path("vignettes", "vutils.R"))
```

```{r}
#| gbp
```

```{r}
#| gv2png
```

# rdecision

<!-- badges: start -->
```{r}
#| results = "asis"
# CRAN status
pander::pandoc.link(
  url = "https://CRAN.R-project.org/package=rdecision",
  text = pander::pandoc.image.return(
    img = "https://www.r-pkg.org/badges/version/rdecision",
    caption = "CRAN status"
  )
)
```

```{r}
#| results = "asis"
# codecov status
pander::pandoc.link(
  url = "https://app.codecov.io/github/ajsims1704/rdecision",
  text = pander::pandoc.image.return(
    img = paste0(
      "https://app.codecov.io/github/ajsims1704/rdecision/branch/master/graph/",
      "badge.svg?token=HHZXK56ZAR"
    ),
    caption = "codecov"
  )
)
```
<!-- badges: end -->

The goal of `rdecision` is to provide methods for assessing health care 
interventions using cohort models (decision trees and semi-Markov models) which
can be constructed using only a few lines of R code. Mechanisms
are provided for associating an uncertainty distribution with each source 
variable and for ensuring transparency of the mathematical relationships between
variables. The package terminology follows Briggs *et al* "Decision Modelling
for Health Economic Evaluation" [@briggs2006].

## Installation

You can install the released version of rdecision from 
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("rdecision")
```

## Examples

```{r}
#nolint start
```

```{r}
library(rdecision)
```

```{r}
#nolint end
```

### A decision tree with parameter uncertainty
```{r}
# read vignette source chunks from corresponding testthat script
knitr::read_chunk(
  file.path("tests", "testthat", "test-model-Lifestyle.R")
)
```

Consider the fictitious and idealized decision problem of choosing between
providing two forms of lifestyle advice, offered to people with vascular
disease, which reduce the risk of needing an interventional procedure. It is
assumed that the interventional procedure is the insertion of a stent, that
the current standard of care is the provision of dietary advice, and that the
new form of care is enrolment on an exercise programme. To assess the decision
problem, we construct a model from the perspective of a healthcare provider,
with a time horizon of one year, and assume that the utility of both forms
of advice is equal. The model evaluates the incremental benefit of the
exercise programme as the incremental number of interventional procedures
avoided against the incremental cost of the exercise programme.

The cost to a healthcare provider of the interventional procedure 
(e.g., inserting a stent) is 5000 GBP; the cost of providing the current form of
lifestyle advice, an appointment with a dietician ("diet"), is 50 GBP and the
cost of providing an alternative form, attendance at an exercise
programme ("exercise"), is 750 GBP. None of the costs are subject to
uncertainty, and are modelled as constant model variables.

```{r}
#| cost-modvars,
#| echo = TRUE
```

If an advice programme is successful, there is no need for an interventional
procedure. In a small trial of the "diet" programme, 12 out of 68 patients
(17.6%) avoided having a procedure, and in a separate small trial of the
"exercise" programme 18 out of 58 patients (31.0%) avoided the procedure. It
is assumed that the baseline characteristics in the two trials were comparable.
The trial results are represented as scalar integers.

```{r}
#| trial-results,
#| echo = TRUE
```

The proportions of the two programmes being successful (i.e., avoiding 
an interventional procedure) are uncertain due to the finite size of each trial
and are represented by model variables with uncertainties which follow Beta
distributions. Probabilities of the failure of the programmes are calculated
using expression model variables to ensure that the total probability
associated with each chance node is one.

```{r}
#| prob-modvars,
#| echo = TRUE
```

The decision tree has one decision node, representing the single choice of the
decision problem (i.e., between the two advice programmes), two chance nodes,
representing whether each programme is a success or failure, and four leaf
nodes (intervention or no intervention for each of the two programmes).

```{r}
#| decision-node,
#| echo = TRUE
```

```{r}
#| chance-nodes,
#| echo = TRUE
```

```{r}
#| leaf-nodes,
#| echo = TRUE
```

There are two action edges emanating from the decision node, which represent the
two choices, and four reaction edges, representing the consequences of the
success and failure of each programme.

```{r}
#| actions,
#| echo = TRUE
```

```{r}
#| reactions,
#| echo = TRUE
```

The decision tree model is constructed from the nodes and edges.

```{r}
#| decision-tree,
#| echo = TRUE
```

```{r}
#| fig.align = "center",
#| include = FALSE
# draw the model and save it in a PNG file within the `man` folder of the
# package, then add a link to it into the README.md script.
pngfile <- file.path("man", "figures", "README-lifestyle.png")
withr::with_png(
  new = pngfile,
  code = {
    dt$draw(border = TRUE)
  }
)
```

```{r}
knitr::include_graphics(path = pngfile)
```

The method `evaluate` is used to calculate the costs and utilities associated
with the decision problem. By default, it evaluates it once with all the model
variables set to their expected values and returns a data frame.

```{r}
#| dt-evaluate-expected,
#| echo = TRUE
```

```{r}
cdiet <- rs[[which(rs[, "Programme"] == "Diet"), "Cost"]]
cexer <- rs[[which(rs[, "Programme"] == "Exercise"), "Cost"]]
delta_c <- cexer - cdiet
```

Examination of the results of evaluation shows that the expected per-patient net
cost of the diet advice programme is `r gbp(cdiet)` GBP and the per-patient net
cost of the exercise programme is `r gbp(cexer)` GBP; i.e., the net cost of the
exercise programme exceeds the diet programme by `r gbp(delta_c)` GBP per
patient. The savings associated with the greater efficacy of the exercise
programme do not offset the increased cost of delivering it.

To estimate the uncertainty of the relative effectiveness, the probabilities of
the success proportions of the two treatments can be represented as model
variables who uncertainty follows a Beta distribution, and the decision tree
re-evaluated.

```{r}
#| actions-probabilistic,
#| echo = TRUE
```

```{r}
#| reactions-probabilistic,
#| echo = TRUE
```

```{r}
#| dt-evaluate-random,
#| echo = TRUE
```
 
```{r}
#| dt-difference
```

The confidence interval of the cost saving is estimated by repeated evaluation
of the tree, each time sampling from the uncertainty distribution of the two
probabilities using, for example, `DT$evaluate(setvars = "random", N = 1000L)`
and inspecting the resulting data frame. From `r N` runs, the 95% confidence
interval of the per patient cost saving 
is `r round(quantile(rs[, "Difference"], probs = 0.025), 2L)` GBP 
to `r round(quantile(rs[, "Difference"], probs = 0.975), 2L)` GBP,
with `r round(100*(sum(rs[, "Difference"] > 0.0) / N), 1L)`% being cost
saving. Although the exercise programme is more costly to provide than the
dietary advice programme, it is more effective and leads to saving overall
because fewer costly interventional procedures are needed. However, due to the
uncertainties of the effectiveness of each programme, it can be concluded that
more evidence is required to be confident that the exercise programme is cost
saving overall.

```{r}
#| threshold
```

The method `threshold` is used to find the threshold of one of the model
variables at which the cost difference reaches zero. By univariate 
threshold analysis, the exercise program will be cost saving when its cost of
delivery is less than `r gbp(cost_threshold)` GBP or when its success rate is
greater than `r gbp(100.0 * success_threshold, 1L)`%. These thresholds are also
subject to uncertainty.

### A three-state Markov model
```{r}
# read vignette source chunks from corresponding testthat script
knitr::read_chunk(
  file.path("tests", "testthat", "test-model-PHV.R")
)
```

Sonnenberg and Beck [-@sonnenberg1993] introduced an illustrative example of a
semi-Markov process with three states: "Well", "Disabled" and "Dead" and one
transition between each state, each with a per-cycle probability. In
`rdecision` such a model is constructed as follows. Note that transitions
from a state to itself must be specified if allowed, otherwise the state would
be a temporary state.

```{r}
#| create-states,
#| echo = TRUE
```

```{r}
#| create-transitions,
#| echo = TRUE
```

```{r}
#| create-model,
#| echo = TRUE
```

```{r}
#| set-pt,
#| echo = TRUE
```

```{r}
#| fig.align = "center",
#| include = FALSE
# represent the model as a graphviz script and render it to an image file using
# the `dot` command line tool, if available, or to a placeholder if not. The
# image file is saved in the `man` folder of the package so it is available to
# the README.md script, to which a link is added to the README.md file.
pngfile <- file.path("man", "figures", "README-phv.png")
gv2png(dot = M$as_DOT(), pngfile = pngfile)
```

```{r}
knitr::include_graphics(path = pngfile)
```

With a starting population of 10,000, the model can be run for 25 years as
follows.
```{r}
#| set-pop,
#| echo = TRUE
```

```{r}
#| cycle,
#| echo = TRUE
```

The output, after rounding, of the `cycles` function is the Markov trace, shown
below, which replicates Table 2 [@sonnenberg1993]. In more recent usage,
cumulative utility is normally called incremental utility, and expressed per
patient (i.e., divided by 10,000).

```{r}
#| trace-to-t2
```

```{r}
keep <- c("Cycle", "Well", "Disabled", "Dead", "CumulativeUtility")
rows <- c(1L, 2L, 3L, 4L, 24L, 25L, 26L)
pander::pander(
  t2[rows, keep], row.names = FALSE, justify = "rrrrr",
  round = c(2L, 0L, 0L, 0L, 3L),
  col.names = c("Years", "Well", "Disabled", "Dead", "Cumulative Utility")
)
```

# Acknowledgements
In addition to using base R [@rcoreteam], `redecision` relies heavily on
the `R6` implementation of classes [@chang2020] and the `rlang` package for
error handling and non-standard evaluation used in expression model variables
[@henry2020]. Building the package vignettes and documentation relies on the
`testthat` package [@wickham2011], the `devtools` package [@wickham2020] and
`rmarkdown` [@xie2018a,@allaire2020,@xie2020].

Underpinning graph theory is based on terminology, definitions and
algorithms from Gross *et al* [-@gross2013], the Wikipedia 
glossary [@wikipedia2021] and links therein. Topological sorting of graphs
is based on Kahn's algorithm [-@kahn1962]. Some of the terminology for decision
trees was based on the work of Kaminski *et al* [-@kaminski2018] and an 
efficient tree drawing algorithm was based on the work of Walker [-@walker1989].
In semi-Markov models, representations are exported in the
DOT language [@gansner1993].

Terminology for decision trees and Markov models in health economic evaluation
was based on the book by Briggs *et al* [-@briggs2006] and the output format
and terminology follows ISPOR recommendations [@briggs2012a,@siebert2012]. 

Citations for examples used in vignettes are given in applicable vignette
files. 

# References

