---
output:
  github_document:
    df_print: kable
    pandoc_args: [
      "--wrap", "preserve"
    ]
bibliography: "vignettes/REFERENCES.bib"
csl: "vignettes/nature-no-et-al.csl"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r}
#| include = FALSE,
#| purl = FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = FALSE,
  comment = "#>"
)
```

# rdecision

```{r}
#| results = "asis"
local({
  writeLines(text = c(
    "<!-- badges: start -->",
    pander::pandoc.link.return(
      url = paste0(
        c("https:/", "cran.r-project.org", "package-rdecision"), collapse = "/"
      ),
      text = pander::pandoc.image.return(
        img = paste0(
          c("https:/", "www.r-pkg.org", "badges", "version", "rdecision"),
          collapse = "/"
        ),
        caption = "CRAN status"
      )
    ),
    pander::pandoc.link.return(
      url = paste0(
        c(
          "https:/", "app.codecov.io", "gh", "ajsims1704",
          "rdecision?branch=master"
        ),
        collapse = "/"
      ),
      text = pander::pandoc.image.return(
        img = paste0(
          c("https:/", "codecov.io", "gh", "ajsims1704", "rdecision", "branch",
            "master", "graph", "badge.svg"), collapse = "/"
        ),
        caption = "Codecov test coverage"
      )
    ),
    "<!-- badges: end -->"
  ))
})
```

The goal of `rdecision` is to provide methods for assessing health care 
interventions using cohort models (decision trees and semi-Markov models) which
can be constructed using only a few lines of R code. Mechanisms
are provided for associating an uncertainty distribution with each source 
variable and for ensuring transparency of the mathematical relationships between
variables. The package terminology follows Briggs *et al* "Decision Modelling
for Health Economic Evaluation" [@briggs2006].

## Installation

You can install the released version of rdecision from 
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("rdecision")
```

## Examples

```{r}
#nolint start
```

```{r}
library(rdecision)
```

```{r}
#nolint end
```

### A decision tree with parameter uncertainty
Consider the fictitious and idealized decision problem of choosing between
providing two forms of lifestyle advice, offered to people with vascular
disease, which reduce the risk of needing an interventional procedure. It is
assumed that the interventional procedure is the insertion of a stent, that
the current standard of care is the provision of dietary advice, and that the
new form of care is enrolment on an exercise programme. To assess the decision
problem, we construct a model from the perspective of a healthcare provider,
with a time horizon of one year, and assume that the utility of both forms
of advice is equal. The model evaluates the incremental benefit of the
exercise programme as the incremental number of interventional procedures
avoided against the incremental cost of the exercise programme.

The cost to a healthcare provider of the interventional procedure 
(e.g., inserting a stent) is 5000 GBP; the cost of providing the current form of
lifestyle advice, an appointment with a dietician ("diet"), is 50 GBP and the
cost of providing an alternative form, attendance at an exercise
programme ("exercise"), is 750 GBP. None of the costs are subject to
uncertainty, and are modelled as constant model variables.

```{r}
#| echo = TRUE
cost_diet <- ConstModVar$new("Cost of diet programme", "GBP", 50.0)
cost_exercise <- ConstModVar$new("Cost of exercise programme", "GBP", 750.0)
cost_stent <- ConstModVar$new("Cost of stent intervention", "GBP", 5000.0)
```

If an advice programme is successful, there is no need for an interventional
procedure. In a small trial of the "diet" programme, 12 out of 68 patients
(17.6%) avoided having a procedure, and in a separate small trial of the
"exercise" programme 18 out of 58 patients (31.0%) avoided the procedure. It
is assumed that the baseline characteristics in the two trials were comparable.
The trial results are represented as scalar integers.

```{r}
#| echo = TRUE
s_diet <- 12L
f_diet <- 56L
s_exercise <- 18L
f_exercise <- 40L
```

The proportions of the two programmes being successful (i.e., avoiding 
an interventional procedure) are uncertain due to the finite size of each trial
and are represented by model variables with uncertainties which follow Beta
distributions. 

```{r}
#| echo = TRUE
p_diet <- BetaModVar$new(
  alpha = s_diet, beta = f_diet, description = "P(diet)", units = ""
)
p_exercise <- BetaModVar$new(
  alpha = s_exercise, beta = f_exercise, description = "P(exercise)", units = ""
)
```

The decision tree has one decision node, representing the single choice of the
decision problem (i.e., between the two advice programmes), two chance nodes,
representing whether each programme is a success or failure, and four leaf
nodes (intervention or no intervention for each of the two programmes).

```{r}
#| echo = TRUE
decision_node <- DecisionNode$new("Programme")
```

```{r}
#| echo = TRUE
chance_node_diet <- ChanceNode$new("Outcome")
chance_node_exercise <- ChanceNode$new("Outcome")
```

```{r}
#| echo = TRUE
leaf_node_diet_no_stent <- LeafNode$new("No intervention")
leaf_node_diet_stent <- LeafNode$new("Intervention")
leaf_node_exercise_no_stent <- LeafNode$new("No intervention")
leaf_node_exercise_stent <- LeafNode$new("Intervention")
```

There are two action edges emanating from the decision node, which represent the
two choices, each with an associated cost.
```{r}
#| echo = TRUE
action_diet <- Action$new(
  decision_node, chance_node_diet, cost = cost_diet, label = "Diet"
)
action_exercise <- Action$new(
  decision_node, chance_node_exercise, cost = cost_exercise, label = "Exercise"
)
```

There are four reaction edges, representing the consequences of the
success and failure of each programme. Edges representing success are associated
with the probability of programmme success, and those representing programme
failure are assigned a probability of `NA` (to ensure that the total probability
associated with each chance node is one) and a failure cost (of fitting a
stent).

```{r}
#| echo = TRUE
reaction_diet_success <- Reaction$new(
  chance_node_diet, leaf_node_diet_no_stent,
  p = p_diet, cost = 0.0, label = "Success"
)

reaction_diet_failure <- Reaction$new(
  chance_node_diet, leaf_node_diet_stent,
  p = NA_real_, cost = cost_stent, label = "Failure"
)

reaction_exercise_success <- Reaction$new(
  chance_node_exercise, leaf_node_exercise_no_stent,
  p = p_exercise, cost = 0.0, label = "Success"
)

reaction_exercise_failure <- Reaction$new(
  chance_node_exercise, leaf_node_exercise_stent,
  p = NA_real_, cost = cost_stent, label = "Failure"
)
```

The decision tree model is constructed from the nodes and edges.

```{r}
#| echo = TRUE
dt <- DecisionTree$new(
  V = list(
    decision_node,
    chance_node_diet,
    chance_node_exercise,
    leaf_node_diet_no_stent,
    leaf_node_diet_stent,
    leaf_node_exercise_no_stent,
    leaf_node_exercise_stent
  ),
  E = list(
    action_diet,
    action_exercise,
    reaction_diet_success,
    reaction_diet_failure,
    reaction_exercise_success,
    reaction_exercise_failure
  )
)
```

```{r}
#| fig.align = "center",
#| include = FALSE
# draw the model and save it in a PNG file within the `man` folder of the
# package, then add a link to it into the README.md script.
pngfile <- file.path("man", "figures", "lifestyle.png")
withr::with_png(
  new = pngfile,
  code = {
    dt$draw()
  }
)
```

```{r}
knitr::include_graphics(path = pngfile)
```

The method `evaluate` is used to calculate the costs and utilities associated
with the decision problem. By default, it evaluates it once with all the model
variables set to their expected values and returns a data frame.

```{r}
#| echo = TRUE
rs <- dt$evaluate()
```

```{r}
cdiet <- rs[[which(rs[, "Programme"] == "Diet"), "Cost"]]
cexer <- rs[[which(rs[, "Programme"] == "Exercise"), "Cost"]]
delta_c <- cexer - cdiet
```

```{r}
#| purl = FALSE
# test that base case evaluation agrees with direct calculation
local({
  # expected costs
  ecdiet <- cost_diet$mean() + (1.0 - p_diet$mean()) * cost_stent$mean()
  ecexer <- cost_exercise$mean() + (1.0 - p_exercise$mean()) * cost_stent$mean()
  # expected cost difference
  ip_diet <- f_diet / (s_diet + f_diet)
  ip_exercise <- f_exercise / (s_exercise + f_exercise)
  nnt <- 1.0 / (ip_diet - ip_exercise)
  incc <- nnt * (cost_exercise$mean() - cost_diet$mean())
  edelta_c <- (incc - cost_stent$mean()) / nnt
  stopifnot(
    all.equal(cdiet, ecdiet, tolerance = 5.0, scale = 1.0),
    all.equal(cexer, ecexer, tolerance = 5.0, scale = 1.0),
    all.equal(delta_c, edelta_c, tolerance = 1.0, scale = 1.0)
  )
})
```

Examination of the results of evaluation shows that the expected per-patient net
cost of the diet advice programme is `r gbp(cdiet)` GBP and the per-patient net
cost of the exercise programme is `r gbp(cexer)` GBP; i.e., the net cost of the
exercise programme exceeds the diet programme by `r gbp(delta_c)` GBP per
patient. The savings associated with the greater efficacy of the exercise
programme do not offset the increased cost of delivering it.

Because the probabilities of the success proportions of the two treatments have
been represented as model variables with an uncertainty distribution, the
uncertainty of the relative effectiveness is estimated by repeated evaluation
of the decision tree.

```{r}
#| echo = TRUE
N <- 1000L
rs <- dt$evaluate(setvars = "random", by = "run", N = N)
```
 
```{r}
rs[, "Difference"] <- rs[, "Cost.Diet"] - rs[, "Cost.Exercise"]
CI <- quantile(rs[, "Difference"], c(0.025, 0.975))
```

The confidence interval of the net cost difference (net cost of the diet
programme minus the net cost of the exercise programme) is estimated from
the resulting data frame. From `r N` runs, the mean net cost difference is
`r round(mean(rs[, "Difference"]), digits = 2L)` GBP with 95% confidence interval
`r round(quantile(rs[, "Difference"], probs = 0.025), 2L)` GBP 
to `r round(quantile(rs[, "Difference"], probs = 0.975), 2L)` GBP,
with `r round(100*(sum(rs[, "Difference"] > 0.0) / N), 1L)`% runs having a lower
net cost for the exercise programme. Although the point estimate net cost of
the exercise programme exceeds that of the diet programme, due to the
uncertainties of the effectiveness of
each programme, it can be concluded that there is insufficient evidence
that the net costs differ.

```{r}
cost_threshold <- dt$threshold(
  index = list(action_exercise),
  ref = list(action_diet),
  outcome = "saving",
  mvd = cost_exercise$description(),
  a = 0.0, b = 5000.0, tol = 0.1
)

success_threshold <- dt$threshold(
  index = list(action_exercise),
  ref = list(action_diet),
  outcome = "saving",
  mvd = p_exercise$description(),
  a = 0.0, b = 1.0, tol = 0.001
)
```

```{r}
#| purl = FALSE
local({
  # number needed to treat
  ip_diet <- f_diet / (s_diet + f_diet)
  ip_exercise <- f_exercise / (s_exercise + f_exercise)
  nnt <- 1.0 / (ip_diet - ip_exercise)
  # expected threshold cost and success probability
  e_cost_threshold <- (cost_stent$mean() / nnt) + cost_diet$mean()
  nnt_threshold <- cost_stent$mean() / (cost_exercise$mean() - cost_diet$mean())
  e_success_threshold <- 1.0 - (ip_diet - (1.0 / nnt_threshold))
  stopifnot(
    all.equal(cost_threshold, e_cost_threshold, tolerance = 5.0, scale = 1.0),
    all.equal(
      success_threshold, e_success_threshold, tolerance = 0.03, scale = 1.0
    )
  )
})
```

The method `threshold` is used to find the threshold of one of the model
variables at which the cost difference reaches zero. By univariate 
threshold analysis, the exercise program will be cost saving when its cost of
delivery is less than `r gbp(cost_threshold)` GBP or when its success rate is
greater than `r gbp(100.0 * success_threshold, 1L)`%. These thresholds are also
subject to uncertainty.

### A three-state Markov model
Sonnenberg and Beck [-@sonnenberg1993] introduced an illustrative example of a
semi-Markov process with three states: "Well", "Disabled" and "Dead" and one
transition between each state, each with a per-cycle probability. In
`rdecision` such a model is constructed as follows. Note that transitions
from a state to itself must be specified if allowed, otherwise the state would
be a temporary state.

```{r}
#| echo = TRUE
# create states
s.well <- MarkovState$new(name = "Well", utility = 1.0)
s.disabled <- MarkovState$new(name = "Disabled", utility = 0.7)
s.dead <- MarkovState$new(name = "Dead", utility = 0.0)
```

```{r}
#| echo = TRUE
# create transitions leaving rates undefined
E <- list(
  Transition$new(s.well, s.well),
  Transition$new(s.dead, s.dead),
  Transition$new(s.disabled, s.disabled),
  Transition$new(s.well, s.disabled),
  Transition$new(s.well, s.dead),
  Transition$new(s.disabled, s.dead)
)
```

```{r}
#| echo = TRUE
# create the model
M <- SemiMarkovModel$new(V = list(s.well, s.disabled, s.dead), E)
```

```{r}
#| echo = TRUE
# create transition probability matrix
snames <- c("Well", "Disabled", "Dead")
Pt <- matrix(
  data = c(0.6, 0.2, 0.2, 0.0, 0.6, 0.4, 0.0, 0.0, 1.0),
  nrow = 3L, byrow = TRUE,
  dimnames = list(source = snames, target = snames)
)
# set the transition rates from per-cycle probabilities
M$set_probabilities(Pt)
```

```{r}
#| include = FALSE
# represent the model as a GML script and render it to an image file using
# the igraph package. The image file is saved in the `man` folder of the package
# so it is available to the README.md script, to which a link is added to the
# README.md file.
pngfile <- file.path("man", "figures", "phv.png")
local({
  # create an igraph object
  gml <- M$as_gml()
  gmlfile <- tempfile(fileext = ".gml")
  writeLines(gml, con = gmlfile)
  ig <- igraph::read_graph(gmlfile, format = "gml")
  # match layout to Sonnenberg and Beck, fig 3
  vxy <- matrix(
    data = c(
      -0.75, +0.75, +0.00,
      +0.75, +0.75, -0.75
    ),
    ncol = 2L,
    dimnames = list(c("Well", "Disabled", "Dead"), c("x", "y"))
  )
  layout <- matrix(
    data = c(
      vapply(X = igraph::V(ig), FUN.VALUE = 1.0, FUN = function(v) {
        lbl <- igraph::vertex_attr(ig, "label", v)
        return(vxy[[lbl, "x"]])
      }),
      vapply(X = igraph::V(ig), FUN.VALUE = 1.0, FUN = function(v) {
        lbl <- igraph::vertex_attr(ig, "label", v)
        return(vxy[[lbl, "y"]])
      })
    ),
    byrow = FALSE,
    ncol = 2L
  )
  # loop angles
  loopa <- vapply(X = igraph::E(ig), FUN.VALUE = 1.0, FUN = function(e) {
    # find source and target labels
    trg <- igraph::head_of(ig, e)
    trgl <- igraph::vertex_attr(ig, name = "label", index = trg)
    src <- igraph::tail_of(ig, e)
    srcl <- igraph::vertex_attr(ig, name = "label", index = src)
    la <- 0.0
    if (trgl == srcl) {
      if (trgl == "Well") {
        la <- pi
      } else if (trgl == "Dead") {
        la <- pi / 2.0
      }
    }
    return(la)
  })
  # plot into png file
  withr::with_png(
    new = pngfile,
    code = {
      withr::with_par(
        new = list(
          oma = c(0L, 0L, 0L, 0L),
          mar = c(3L, 3L, 3L, 3L),
          xpd = NA
        ),
        code = {
          plot(
            ig,
            rescale = FALSE, asp = 0L,
            vertex.shape = "circle", vertex.size = 60.0,
            vertex.color = "white", vertex.label.color = "black",
            edge.color = "black",
            edge.arrow.size = 0.75,
            frame = FALSE,
            layout = layout,
            loop.size = 0.8,
            edge.loop.angle = loopa
          )
        }
      )
    }
  )
})
```

```{r}
#| fig.align = "center",
knitr::include_graphics(path = pngfile)
```

With a starting population of 10,000, the model can be run for 24 years as
follows.
```{r}
#| echo = TRUE
# set the starting populations
M$reset(c(Well = 10000.0, Disabled = 0.0, Dead = 0.0))
```

```{r}
#| echo = TRUE
# cycle
MT <- M$cycles(24L, hcc.pop = FALSE, hcc.cost = FALSE, hcc.QALY = FALSE)
```

The output, after rounding, of the `cycles` function is the Markov trace, shown
below, which replicates Table 2 [@sonnenberg1993]. In more recent usage,
cumulative utility is normally called incremental utility, and expressed per
patient (i.e., divided by 10,000).

```{r}
t2 <- data.frame(
  Cycle = MT[, "Cycle"],
  Well = round(MT[, "Well"], 0L),
  Disabled = round(MT[, "Disabled"], 0L),
  Dead = round(MT[, "Dead"], 0L),
  CycleSum = round(MT[, "QALY"] * 10000.0, 0L),
  CumulativeUtility = round(10000.0 * cumsum(MT[, "QALY"]), 0L)
)
```

```{r}
#| purl = FALSE
# test that reformatted cycle results match S&B table 2
local({
  # cycle 0
  r0 <- which(t2[, "Cycle"] == 0L)
  stopifnot(
    all.equal(t2[[r0, "Well"]], 10000.0),
    all.equal(t2[[r0, "Disabled"]], 0.0),
    all.equal(t2[[r0, "Dead"]], 0.0),
    all.equal(t2[[r0, "CycleSum"]], 0.0),
    all.equal(t2[[r0, "CumulativeUtility"]], 0.0)
  )
  # cycle 1
  r <- which(t2[, "Cycle"] == 1L)
  stopifnot(
    all.equal(t2[[r, "Well"]], 6000.0),
    all.equal(t2[[r, "Disabled"]], 2000.0),
    all.equal(t2[[r, "Dead"]], 2000.0),
    all.equal(
      t2[[r, "CycleSum"]], 7400.0, tolerance = 1.0, scale = 1.0
    ),
    all.equal(
      t2[[r, "CumulativeUtility"]], 7400.0,
      tolerance = 1.0, scale = 1.0
    )
  )
  # cycle 2
  r <- which(t2[, "Cycle"] == 2L)
  stopifnot(
    all.equal(t2[[r, "Well"]], 3600.0),
    all.equal(t2[[r, "Disabled"]], 2400.0),
    all.equal(t2[[r, "Dead"]], 4000.0),
    all.equal(
      t2[[r, "CycleSum"]], 5280.0, tolerance = 1.0, scale = 1.0
    ),
    all.equal(
      t2[[r, "CumulativeUtility"]], 12680.0,
      tolerance = 1.0, scale = 1.0
    )
  )
  # cycle 23
  r <- which(t2[, "Cycle"] == 23L)
  stopifnot(
    all.equal(t2[[r, "Well"]], 0.0, tolerance = 1.0, scale = 1.0),
    all.equal(t2[[r, "Disabled"]], 1.0, tolerance = 1.0, scale = 1.0),
    all.equal(t2[[r, "Dead"]], 9999.0, tolerance = 1.0, scale = 1.0),
    all.equal(t2[[r, "CycleSum"]], 0.0, tolerance = 1.0, scale = 1.0),
    all.equal(
      t2[[r, "CumulativeUtility"]], 23752.0,
      tolerance = 5.0, scale = 1.0
    )
  )
  # cycle 24
  r <- which(t2[, "Cycle"] == 24L)
  stopifnot(
    all.equal(t2[[r, "Well"]], 0.0, tolerance = 1.0, scale = 1.0),
    all.equal(t2[[r, "Disabled"]], 0.0, tolerance = 1.0, scale = 1.0),
    all.equal(t2[[r, "Dead"]], 10000.0, tolerance = 1.0, scale = 1.0),
    all.equal(t2[[r, "CycleSum"]], 0.0, tolerance = 1.0, scale = 1.0),
    all.equal(
      t2[[r, "CumulativeUtility"]], 23752.0,
      tolerance = 5.0, scale = 1.0
    )
  )
})
```

```{r}
keep <- c("Cycle", "Well", "Disabled", "Dead", "CumulativeUtility")
rows <- c(1L, 2L, 3L, 4L, 24L, 25L)
with(data = t2[rows, keep], expr = {
  data.frame(
    Years = round(Cycle, digits = 2L),
    Well = round(Well, digits = 0L),
    Disabled = round(Disabled, digits = 0L),
    "Cumulative Utility" = round(CumulativeUtility, digits = 3L),
    stringsAsFactors = FALSE
  )
})
```

# Acknowledgements
In addition to using base R [@rcoreteam], `redecision` relies on
the `R6` implementation of classes [@chang2020] and the `rlang` package for
error handling and non-standard evaluation used in expression model variables
[@henry2020]. Building the package vignettes and documentation relies on the
`testthat` package [@wickham2011], the `devtools` package [@wickham2020], 
`rmarkdown` [@xie2018a,@allaire2020,@xie2020] and `knitr` [@xie2024].

Underpinning graph theory is based on terminology, definitions and
algorithms from Gross *et al* [-@gross2013], the Wikipedia 
glossary [@wikipedia2021] and links therein. Topological sorting of graphs
is based on Kahn's algorithm [-@kahn1962]. Some of the terminology for decision
trees was based on the work of Kaminski *et al* [-@kaminski2018] and an 
efficient tree drawing algorithm was based on the work of Walker [-@walker1989].
In semi-Markov models, representations are exported in the
DOT language [@gansner1993].

Terminology for decision trees and Markov models in health economic evaluation
was based on the book by Briggs *et al* [-@briggs2006] and the output format
and terminology follows ISPOR recommendations [@briggs2012a,@siebert2012]. 

Citations for examples used in vignettes are given in applicable vignette
files. 

# References

