---
title: "Elementary decision tree (Evans 1997)"
subtitle: "Sumatriptan versus caffeine for migraine"
author: "Andrew J. Sims"
date: "April 2020"
bibliography: "REFERENCES.bib"
csl: "nature-no-et-al.csl"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_caption: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Elementary decision tree (Evans 1997)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include = FALSE,
#| purl = FALSE
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| purl = FALSE
#nolint start
```

```{r}
library(rdecision)
```

```{r}
#| purl = FALSE
#nolint end
```

# Introduction
This vignette is an example of modelling a decision tree using the `rdecision`
package. It is based on the example given by Briggs [-@briggs2006] (Box 2.3) 
which itself is based on a decision tree which compared oral Sumatriptan versus
oral caffeine/Ergotamine for migraine [@evans1997]. In this vignette, we 
consider the problem from the perspective of a provincial health department.

# Creating the model

## Model variables
The following code defines the variables for cost, utility and effect that will
be used in the model. There are 14 variables in total; 4 costs, 4 utilities and
6 probabilities.
```{r}
#| echo = TRUE
# Time horizon
th <- as.difftime(24L, units = "hours")

# model variables for cost
c_sumatriptan <- 16.10
c_caffeine <- 1.32
c_ed <- 63.16
c_admission <- 1093.0

# model variables for utility
u_relief_norecurrence <- 1.0
u_relief_recurrence <- 0.9
u_norelief_endures <- -0.30
u_norelief_er <- 0.1

# model variables for effect
p_sumatriptan_recurrence <- 0.594
p_caffeine_recurrence <- 0.703
p_sumatriptan_relief <- 0.558
p_caffeine_relief <- 0.379
p_er <- 0.08
p_admitted <- 0.002
```

## Constructing the tree
The following code constructs the decision tree. In the
formulation used by `rdecision`, a decision tree is a form of
*arborescence*, a directed graph of nodes and edges, with a single
root and a unique path from the root to each leaf node. Decision trees
comprise three types of node: decision, chance and leaf nodes and two
types of edge: actions (whose sources are decision nodes) and reactions
(whose sources are chance nodes), [see Figure 1](#tree-diagram).
If the probability of traversing one reaction edge from any chance node is set
to `NA_real_`, it will be calculated as 1 minus the sum of probabilities of the
other reaction edges from that node when the tree is evaluated.

```{r}
#| echo = TRUE
# Sumatriptan branch
ta <- LeafNode$new("A", utility = u_relief_norecurrence, interval = th)
tb <- LeafNode$new("B", utility = u_relief_recurrence, interval = th)
c3 <- ChanceNode$new()
e1 <- Reaction$new(
  c3, ta, p = p_sumatriptan_recurrence, label = "No recurrence"
)
e2 <- Reaction$new(
  c3, tb, p = NA_real_, cost = c_sumatriptan, label = "Relieved 2nd dose"
)
td <- LeafNode$new("D", utility = u_norelief_er, interval = th)
te <- LeafNode$new("E", utility = u_norelief_endures, interval = th)
c7 <- ChanceNode$new()
e3 <- Reaction$new(c7, td, p = NA_real_, label = "Relief")
e4 <- Reaction$new(
  c7, te, p = p_admitted, cost = c_admission, label = "Hospitalization"
)

tc <- LeafNode$new("C", utility = u_norelief_endures, interval = th)
c4 <- ChanceNode$new()
e5 <- Reaction$new(c4, tc, p = NA_real_, label = "Endures attack")
e6 <- Reaction$new(c4, c7, p = p_er, cost = c_ed, label = "ER")

c1 <- ChanceNode$new()
e7 <- Reaction$new(c1, c3, p = p_sumatriptan_relief, label = "Relief")
e8 <- Reaction$new(c1, c4, p = NA_real_, label = "No relief")

# Caffeine/Ergotamine branch
tf <- LeafNode$new("F", utility = u_relief_norecurrence, interval = th)
tg <- LeafNode$new("G", utility = u_relief_recurrence, interval = th)
c5 <- ChanceNode$new()
e9 <- Reaction$new(c5, tf, p = p_caffeine_recurrence, label = "No recurrence")
e10 <- Reaction$new(
  c5, tg, p = NA_real_, cost = c_caffeine, label = "Relieved 2nd dose"
)
ti <- LeafNode$new("I", utility = u_norelief_er, interval = th)
tj <- LeafNode$new("J", utility = u_norelief_endures, interval = th)
c8 <- ChanceNode$new()
e11 <- Reaction$new(c8, ti, p = NA_real_, label = "Relief")
e12 <- Reaction$new(
  c8, tj, p = p_admitted, cost = c_admission, label = "Hospitalization"
)

th <- LeafNode$new("H", utility = u_norelief_endures, interval = th)
c6 <- ChanceNode$new()
e13 <- Reaction$new(c6, th, p = NA_real_, label = "Endures attack")
e14 <- Reaction$new(c6, c8, p = p_er, cost = c_ed, label = "ER")

c2 <- ChanceNode$new()
e15 <- Reaction$new(c2, c5, p = p_caffeine_relief, label = "Relief")
e16 <- Reaction$new(c2, c6, p = NA_real_, label = "No relief")

# decision node
d1 <- DecisionNode$new("d1")
e17 <- Action$new(d1, c1, cost = c_sumatriptan, label = "Sumatriptan")
e18 <- Action$new(d1, c2, cost = c_caffeine, label = "Caffeine-Ergotamine")

# create lists of nodes and edges
V <- list(
  d1, c1, c2, c3, c4, c5, c6, c7, c8,
  ta, tb, tc, td, te, tf, tg, th, ti, tj
)
E <- list(
  e1, e2, e3, e4, e5, e6, e7, e8, e9, e10, e11, e12, e13, e14, e15, e16,
  e17, e18
)

# tree
dt <- DecisionTree$new(V, E)
```

```{r}
#| purl = FALSE
# test that decision tree structure is as per Evans et al
stopifnot(
  all.equal(d1$label(), "d1")
)
```

```{r}
#| results = "hide",
#| fig.keep = "all",
#| fig.align = "center",
#| fig.cap = "Figure 1. Decision tree for the Sumatriptan model"
dt$draw(border = TRUE)
```

# Running the model
The method `evaluate` of decision tree objects computes
the probability, cost and utility of each *strategy* for the model. A strategy
is a unanimous prescription of the actions at each decision node. In this
example there is a single decision node with two actions, and the strategies
are simply the two forms of treatment to be compared. More complex decision
trees are also possible.

The paths traversed in each strategy can be evaluated individually
using the method `evaluate(by = "path")`. In `rdecision` a strategy is defined
as a set of action edges with one action edge per decision node. It is necessary
to use the option `by = "path"` only if information about each pathway is
required; normally it is sufficient to call `evaluate` which will automatically
aggregate the evaluation by strategy.

# Model results

## Base case
The evaluation of each pathway, for each strategy, is done as follows:
```{r}
#| echo = TRUE
ep <- dt$evaluate(by = "path")
```

```{r}
#| purl = FALSE
# test that evaluation by path is as per Box 2.3 of Briggs
local({
  stopifnot(
    all.equal(nrow(ep), 10L),
    setequal(
      colnames(ep),
      c(
        "Leaf", "d1", "Probability", "Cost", "Benefit", "Utility", "QALY",
        "Run"
      )
    ),
    setequal(ep[, "Leaf"], LETTERS[1L : 10L]),
    all.equal(sum(ep[, "Probability"]), 2.0, tolerace = 0.01, scale = 1.0)
  )
  ia <- which(ep[, "Leaf"] == "A")
  stopifnot(
    all.equal(ep[[ia, "d1"]], "Sumatriptan"),
    all.equal(ep[[ia, "Probability"]], 0.331, tolerance = 0.001, scale = 1.0),
    all.equal(ep[[ia, "Cost"]], 5.34, tolerance = 0.01, scale = 1.0),
    all.equal(ep[[ia, "Utility"]], 0.33, tolerance = 0.01, scale = 1.0)
  )
  ih <- which(ep[, "Leaf"] == "H")
  stopifnot(
    all.equal(ep[[ih, "d1"]], "Caffeine-Ergotamine"),
    all.equal(ep[[ih, "Probability"]], 0.571, tolerance = 0.001, scale = 1.0),
    all.equal(ep[[ih, "Cost"]], 0.75, tolerance = 0.01, scale = 1.0),
    all.equal(ep[[ih, "Utility"]], -0.17, tolerance = 0.01, scale = 1.0)
  )
})
```

and yields the following table:
```{r}
#| echo = FALSE
with(data = ep, expr = {
  data.frame(
    Leaf = Leaf,
    Probability = round(Probability, digits = 4L),
    Cost = round(Cost, digits = 2L),
    Utility = round(Utility, digits = 5L),
    stringsAsFactors = FALSE
  )
})
```

There are, as expected, ten pathways (5 per strategy). The expected
cost, utility and QALY (utility multiplied by the time horizon of the model) for
each choice can be calculated from the table 
above, or by invoking the `evaluate` method of a decision tree object with the
default parameter `by = "strategy"`. 

```{r}
#| echo = TRUE
es <- dt$evaluate()
```

This gives the following result, consistent with that reported by
Evans *et al* [-@evans1997].

```{r}
#| echo = FALSE
with(data = es, expr = {
  data.frame(
    d1 = d1,
    Cost = round(Cost, digits = 2L),
    Utility = round(Utility, digits = 4L),
    QALY = round(QALY, digits = 4L),
    stringsAsFactors = FALSE
  )
})
```

```{r}
#| echo = FALSE
is <- which(es[, "d1"] == "Sumatriptan")
cost_s <- es[[is, "Cost"]]
utility_s <- es[[is, "Utility"]]
qaly_s <- es[[is, "QALY"]]

ic <- which(es[, "d1"] == "Caffeine-Ergotamine")
cost_c <- es[[ic, "Cost"]]
utility_c <- es[[ic, "Utility"]]
qaly_c <- es[[ic, "QALY"]]

delta_c <- cost_s - cost_c
delta_u <- utility_s - utility_c
delta_q <- qaly_s - qaly_c
icer <- delta_c / delta_q
```

```{r}
#| purl = FALSE
# test that evaluation by strategy is as per Evans et al
stopifnot(
  all.equal(nrow(es), 2L),
  setequal(
    colnames(es),
    c("d1", "Run", "Probability", "Cost", "Benefit", "Utility", "QALY")
  ),
  setequal(es[, "d1"], c("Sumatriptan", "Caffeine-Ergotamine")),
  all.equal(sum(es["Probability"]), 2.0, tolerance = 0.01, scale = 1.0),
  all.equal(cost_s, 22.06, tolerance = 0.01, scale = 1.0),
  all.equal(utility_s, 0.41, tolerance = 0.01, scale = 1.0),
  all.equal(cost_c, 4.73, tolerance = 0.02, scale = 1.0),
  all.equal(utility_c, 0.20, tolerance = 0.01, scale = 1.0),
  icer / 29366.0 >= 0.95,
  icer / 29366.0 <= 1.05
)
```

The incremental cost was $Can `r gbp(x = delta_c, p = TRUE)`
(`r gbp(x = cost_s, p = TRUE)` - `r gbp(x = cost_c, p = TRUE)`)
and the incremental utility was `r round(delta_u, 2L)`
(`r round(utility_s, 2L)` - `r round(utility_c, 2L)`). Because the time
horizon of the model was 1 day, the incremental QALYs was the incremental
annual utility divided by 365, and the ICER was therefore equal to `r gbp(icer)`
\$Can/QALY, within 5% of the published estimate (29,366 \$Can/QALY). 

## Univariate sensitivity analysis
Evans *et al* [-@evans1997] reported the ICER for various alternative values
of input variables. For example (their Table VIII), they reported that the 
ICER was 60,839 $Can/QALY for a relative increase in effectiveness of 9.1%
(i.e., when the relief from Sumatriptan was 9.1 percentage points greater than
that of Caffeine-Ergotamine) and 18,950 $Can/QALY for a relative increase in
effectiveness of 26.8% (these being the lower and upper confidence intervals
of the estimate of effectiveness from meta-analysis).

To calculate these ICERs, we set the value of the model variable
`p_sumatriptan_relief`, and re-evaluate the model. The lower range of ICER
(with the greater relative increase in effectiveness) is calculated as follows:

```{r}
#| echo = TRUE
p_sumatriptan_relief <- p_caffeine_relief + 0.268
e7$set_probability(p_sumatriptan_relief)
es <- dt$evaluate()
```

```{r}
#| echo = FALSE
is <- which(es[, "d1"] == "Sumatriptan")
cost_s_upper <- es[[is, "Cost"]]
utility_s_upper <- es[[is, "Utility"]]
qaly_s_upper <- es[[is, "QALY"]]

ic <- which(es[, "d1"] == "Caffeine-Ergotamine")
cost_c_upper <- es[[ic, "Cost"]]
utility_c_upper <- es[[ic, "Utility"]]
qaly_c_upper <- es[[ic, "QALY"]]

delta_c_upper <- cost_s_upper - cost_c_upper
delta_u_upper <- utility_s_upper - utility_c_upper
delta_q_upper <- qaly_s_upper - qaly_c_upper
icer_upper <- delta_c_upper / delta_q_upper
```

```{r}
#| purl = FALSE
# test that upper relief threshold ICER agrees with Evans et al
stopifnot(
  icer_upper / 18950.0 >= 0.95,
  icer_upper / 18950.0 <= 1.05
)
```

This yields the following table, from which the ICER is calculated as 
`r gbp(icer_upper)` \$Can/QALY, close to the published estimate of
18,950 \$Can/QALY.

```{r}
#| echo = FALSE
with(data = es, expr = {
  data.frame(
    d1 = d1,
    Cost = round(Cost, digits = 2L),
    Utility = round(Utility, digits = 4L),
    QALY = round(QALY, digits = 4L),
    stringsAsFactors = FALSE
  )
})
```

The upper range of ICER (with the smaller relative increase in effectiveness) is
calculated as follows:

```{r}
#| echo = TRUE
p_sumatriptan_relief <- p_caffeine_relief + 0.091
e7$set_probability(p_sumatriptan_relief)
es <- dt$evaluate()
```

```{r}
#| echo = FALSE
is <- which(es[, "d1"] == "Sumatriptan")
cost_s_lower <- es[[is, "Cost"]]
utility_s_lower <- es[[is, "Utility"]]
qaly_s_lower <- es[[is, "QALY"]]

ic <- which(es[, "d1"] == "Caffeine-Ergotamine")
cost_c_lower <- es[[ic, "Cost"]]
utility_c_lower <- es[[ic, "Utility"]]
qaly_c_lower <- es[[ic, "QALY"]]

delta_c_lower <- cost_s_lower - cost_c_lower
delta_u_lower <- utility_s_lower - utility_c_lower
delta_q_lower <- qaly_s_lower - qaly_c_lower
icer_lower <- delta_c_lower / delta_q_lower
```

```{r}
#| purl = FALSE
# test that lower relief threshold ICER agrees with Evans et al
stopifnot(
  icer_lower / 60839.0 >= 0.95,
  icer_lower / 60839.0 <= 1.05
)
```

This yields the following table, from which the ICER is calculated as 
`r gbp(icer_lower)` \$Can/QALY, close to the published estimate of
60,839 \$Can/QALY.

```{r}
#| echo = FALSE
with(data = es, expr = {
  data.frame(
    d1 = d1,
    Cost = round(Cost, digits = 2L),
    Utility = round(Utility, digits = 4L),
    QALY = round(QALY, digits = 4L),
    stringsAsFactors = FALSE
  )
})
```

```{r}
#| purl = FALSE
# test that upper and lower ICER thresholds can be replicatd with thresholding
local({

  # model variables with uncertainty
  p_sumatriptan_relief <- ConstModVar$new(
    "P(relief|sumatriptan)", "P", 0.558
  )

  # set probabilities for edges associated with model variables
  e7$set_probability(p_sumatriptan_relief)
  e15$set_probability(p_caffeine_relief)

  # upper 95% relief rate threshold for ICER (Table VIII)
  p_relief_upper <- dt$threshold(
    index = list(e17), ref = list(e18), outcome = "ICER",
    mvd = p_sumatriptan_relief$description(),
    a = 0.6, b = 0.7,
    lambda = 18950.0, tol = 0.0001
  )
  # lower 95% relief rate threshold for ICER (Table VIII)
  p_relief_lower <- dt$threshold(
    index = list(e17), ref = list(e18), outcome = "ICER",
    mvd = p_sumatriptan_relief$description(),
    a = 0.4, b = 0.5,
    lambda = 60839.0, tol = 0.0001
  )

  # check parameters of threshold function
  # mean relief rate threshold for ICER
  pt <- dt$threshold(
    index = list(e17), ref = list(e18), outcome = "ICER",
    mvd = p_sumatriptan_relief$description(),
    a = 0.5, b = 0.6,
    lambda = 29366.0, tol = 0.0001
  )
  # check values against Table VIII
  stopifnot(
    all.equal(pt, p_caffeine_relief + 0.179, tolerance = 0.02, scale = 1.0),
    all.equal(
      p_relief_upper, p_caffeine_relief + 0.268, tolerance = 0.02, scale = 1.0
    ),
    all.equal(
      p_relief_lower, p_caffeine_relief + 0.091, tolerance = 0.02, scale = 1.0
    )
  )
})
```

```{r}
#| purl = FALSE
# test that ICERs computed by tornado function are as expected
local({

  # probability variables with uncertainty
  p_sumatriptan_relief <- ConstModVar$new(
    "P(relief|sumatriptan)", "P", 0.558
  )
  e7$set_probability(p_sumatriptan_relief)
  e15$set_probability(p_caffeine_relief)

  # cost variables with uncertainty
  c_sumatriptan <- GammaModVar$new(
    "Sumatriptan", "CAD", shape = 16.10, scale = 1.0
  )
  c_caffeine <- GammaModVar$new(
    "Caffeine", "CAD", shape = 1.32, scale = 1.0
  )
  e2$set_cost(c_sumatriptan)
  e10$set_cost(c_caffeine)
  e17$set_cost(c_sumatriptan)
  e18$set_cost(c_caffeine)

  # check ICER ranges in tornado diagram (branches B and G get 2nd dose)
  TO <- dt$tornado(
    index = list(e17), ref = list(e18), outcome = "ICER", draw = FALSE
  )
  c_sumatriptan$set("expected")
  c_caffeine$set("expected")
  x <- qgamma(p = 0.025, shape = 16.10, rate = 1.0)
  deltac <- (x - c_sumatriptan$get()) * 1.227
  stopifnot(
    all.equal(
      TO[[which(TO$Description == "Sumatriptan"), "LL"]],
      x,
      tolerance = 0.01,
      scale = 1.0
    ),
    all.equal(
      TO[[which(TO$Description == "Sumatriptan"), "outcome.min"]],
      (cost_s - cost_c + deltac) / delta_q,
      tolerance = 100.0,
      scale = 1.0
    )
  )
  x <- qgamma(p = 0.975, shape = 16.10, rate = 1.0)
  deltac <- (x - c_sumatriptan$get()) * 1.227
  stopifnot(
    all.equal(
      TO[[which(TO$Description == "Sumatriptan"), "UL"]],
      x,
      tolerance = 0.01,
      scale = 1.0
    ),
    all.equal(
      TO[[which(TO$Description == "Sumatriptan"), "outcome.max"]],
      (cost_s - cost_c + deltac) / delta_q,
      tolerance = 100.0,
      scale = 1.0
    )
  )
  x <- qgamma(p = 0.025, shape = 1.32, rate = 1.0)
  deltac <- (c_caffeine$get() - x) * 1.113
  stopifnot(
    all.equal(
      TO[[which(TO$Description == "Caffeine"), "LL"]],
      x,
      tolerance = 0.01,
      scale = 1.0
    ),
    all.equal(
      TO[[which(TO$Description == "Caffeine"), "outcome.min"]],
      (cost_s - cost_c + deltac) / delta_q,
      tolerance = 100.0,
      scale = 1.0
    )
  )
  x <- qgamma(p = 0.975, shape = 1.32, rate = 1.0)
  deltac <- (c_caffeine$get() - x) * 1.113
  stopifnot(
    all.equal(
      TO[[which(TO$Description == "Caffeine"), "UL"]],
      x,
      tolerance = 0.01,
      scale = 1.0
    ),
    all.equal(
      TO[[which(TO$Description == "Caffeine"), "outcome.max"]],
      (cost_s - cost_c + deltac) / delta_q,
      tolerance = 100.0,
      scale = 1.0
    )
  )
})
```

# References
