---
title: "Elementary decision tree (Evans 1997)"
subtitle: "Sumatriptan versus caffeine for migraine"
author: "Andrew J. Sims"
date: "April 2020"
bibliography: "REFERENCES.bib"
csl: "nature-no-et-al.csl"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_caption: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Elementary decision tree (Evans 1997)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| purl = FALSE,
#| include = FALSE
# read vignette source chunks from corresponding testthat script
knitr::read_chunk(
  file.path("..", "tests", "testthat", "test-model-Sumatriptan.R")
)
```

```{r}
#| include = FALSE,
#| purl = FALSE
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| purl = FALSE
#nolint start
```

```{r}
library(rdecision)
```

```{r}
#| purl = FALSE
#nolint end
```

# Introduction
This vignette is an example of modelling a decision tree using the `rdecision`
package. It is based on the example given by Briggs [-@briggs2006] (Box 2.3) 
which itself is based on a decision tree which compared oral Sumatriptan versus
oral caffeine/Ergotamine for migraine [@evans1997]. In this vignette, we 
consider the problem from the perspective of a provincial health department.

# Creating the model

## Model variables
The following code defines the variables for cost, utility and effect that will
be used in the model. There are 14 variables in total; 4 costs, 4 utilities and
6 probabilities.
```{r}
#| modvars,
#| echo = TRUE
```

## Constructing the tree
The following code constructs the decision tree. In the
formulation used by `rdecision`, a decision tree is a form of
*arborescence*, a directed graph of nodes and edges, with a single
root and a unique path from the root to each leaf node. Decision trees
comprise three types of node: decision, chance and leaf nodes and two
types of edge: actions (whose sources are decision nodes) and reactions
(whose sources are chance nodes), [see Figure 1](#tree-diagram).

```{r}
#| model,
#| echo = TRUE
```

```{r}
#| results = "hide",
#| fig.keep = "all",
#| fig.align = "center",
#| fig.cap = "Figure 1. Decision tree for the Sumatriptan model"
DT$draw(border = TRUE)
```

# Running the model
The method `evaluate` of decision tree objects computes
the probability, cost and utility of each *strategy* for the model. A strategy
is a unanimous prescription of the actions at each decision node. In this
example there is a single decision node with two actions, and the strategies
are simply the two forms of treatment to be compared. More complex decision
trees are also possible.

The paths traversed in each strategy can be evaluated individually
using the method `evaluate(by = "path")`. In `rdecision` a strategy is defined
as a set of action edges with one action edge per decision node. It is necessary
to use the option `by = "path"` only if information about each pathway is
required; normally it is sufficient to call `evaluate` which will automatically
aggregate the evaluation by strategy.

# Model results

## Base case
The evaluation of each pathway, for each strategy, is done as follows:
```{r}
#| eval-by-path,
#| echo = TRUE
```

and yields the following table:
```{r}
#| echo = FALSE
knitr::kable(
  ep[, c("Leaf", "Probability", "Cost", "Utility")],
  align = "lrrr",
  digits = c(2L, 4L, 2L, 5L),
  format.args = list(scientific = FALSE)
)
```

There are, as expected, ten pathways (5 per strategy). The expected
cost, utility and QALY (utility multiplied by the time horizon of the model) for
each choice can be calculated from the table 
above, or by invoking the `evaluate` method of a decision tree object with the
default parameter `by = "strategy"`. 

```{r}
#| eval-by-strategy,
#| echo = TRUE
```

This gives the following result, consistent with that reported by
Evans *et al* [-@evans1997].

```{r}
#| echo = FALSE
knitr::kable(
  es[, c("d1", "Cost", "Utility", "QALY")],
  align = "lrrr",
  digits = c(2L, 2L, 4L, 4L),
  format.args = list(scientific = FALSE)
)
```

```{r}
#| icer-basecase,
#| echo = FALSE
```
The incremental cost was $Can `r gbp(x = delta_c, p = TRUE)`
(`r gbp(x = cost_s, p = TRUE)` - `r gbp(x = cost_c, p = TRUE)`)
and the incremental utility was `r round(delta_u, 2L)`
(`r round(utility_s, 2L)` - `r round(utility_c, 2L)`). Because the time
horizon of the model was 1 day, the incremental QALYs was the incremental
annual utility divided by 365, and the ICER was therefore equal to `r gbp(icer)`
\$Can/QALY, within 5% of the published estimate (29,366 \$Can/QALY). 

## Univariate sensitivity analysis
Evans *et al* [-@evans1997] reported the ICER for various alternative values
of input variables. For example (their Table VIII), they reported that the 
ICER was 60,839 $Can/QALY for a relative increase in effectiveness of 9.1%
(i.e., when the relief from Sumatriptan was 9.1 percentage points greater than
that of Caffeine-Ergotamine) and 18,950 $Can/QALY for a relative increase in
effectiveness of 26.8% (these being the lower and upper confidence intervals
of the estimate of effectiveness from meta-analysis).

To calculate these ICERs, we set the value of the model variable
`p_sumatriptan_relief`, and re-evaluate the model. The lower range of ICER
(with the greater relative increase in effectiveness) is calculated as follows:

```{r}
#| relief-threshold-upper,
#| echo = TRUE
```
```{r}
#| icer-upper,
#| echo = FALSE
```
This yields the following table, from which the ICER is calculated as 
`r gbp(icer_upper)` \$Can/QALY, close to the published estimate of
18,950 \$Can/QALY.

```{r}
#| echo = FALSE
knitr::kable(
  es[, c("d1", "Cost", "Utility", "QALY")],
  align = "lrrr",
  digits = c(2L, 2L, 4L, 4L),
  format.args = list(scientific = FALSE)
)
```

The upper range of ICER (with the smaller relative increase in effectiveness) is
calculated as follows:

```{r}
#| relief-threshold-lower,
#| echo = TRUE
```
```{r}
#| icer-lower,
#| echo = FALSE
```
This yields the following table, from which the ICER is calculated as 
`r gbp(icer_lower)` \$Can/QALY, close to the published estimate of
60,839 \$Can/QALY.

```{r}
#| echo = FALSE
knitr::kable(
  es[, c("d1", "Cost", "Utility", "QALY")],
  align = "lrrr",
  digits = c(2L, 2L, 4L, 4L),
  format.args = list(scientific = FALSE)
)
```

# References
