---
title: "Tegaderm CHG IV Securement Dressing for Central Venous and Arterial Catheter Insertion Sites"
subtitle: "A decision tree example with probabilistic sensitivity analysis"
author: "Andrew J. Sims"
date: "July 2020"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Tegaderm CHG IV Securement Dressing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "references.bib"
csl: "bmj.csl"
---
```{r, include = FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE, echo=FALSE}
library("rdecision")
library("rlang")
library("pander")
```

# Introduction
This vignette is an example of modelling a decision tree using the `rdecision`
package, with probabilistic sensitivity analysis. It is based on the model 
reported by Jenks *et al* [-@jenks2016] in which a transparent dressing used 
to secure vascular catheters (Tegaderm CHG) was compared with a 
standard dressing.

# Model variables
Thirteen variables were used in the model. The choice of
variables, their distributions and their parameters are taken from table 3 
of Jenks *et al* [-@jenks2016], with the following additional information:

* For variables with lognormal uncertainty, the manufacturer synthesized a log
  normal distribution as $\exp(\alpha + \beta r())$ where $r()$ is a random
  draw from a standard normal distribution. This does not follow the 
  parametrizations of the LogNormModVar provided in `rdecision`, but can
  be reproduced using expression model variables. Three standard normal
  model variables were introduced for this purpose.
* The values for
  $\alpha$ and $\beta$ for the hazard ratio for CRBSI and LSI were
  -0.911 and -0.393 respectively, and 1.482 and 0.490 for the relative
  risk of dermatitis. 
* The probabilities of CRBSI and LSI for standard dressings ($p$) were 
  modified by the hazard ratio for Tegaderm using the form $(1-(1-p)^h)$
  where $h$ is the hazard ratio. Relative risks were applied as
  multipliers.
* The point estimate cost of CRBSI was £9900, not £9990, although the parameters
  (198,50) are quoted correctly.

The model variables were constructed as follows:

```{r variables, echo=T}
# standard normals
n1 <- NormModVar$new("SN1","", 0, 1)
n2 <- NormModVar$new("SN2","", 0, 1)
n3 <- NormModVar$new("SN3","", 0, 1)
  
# clinical variables
r.CRBSI <- NormModVar$new(
  'Baseline CRBSI rate', '/1000 catheter days', mu=1.48, sigma=0.074
)
hr.CRBSI <- ExprModVar$new(
  "Tegaderm CRBSI HR", 
  "HR", 
  rlang::quo(exp(-0.911-0.393*n1))
)
hr.LSI <- ExprModVar$new(
  "Tegaderm LSI HR", 
  "HR", 
  rlang::quo(exp(-0.911-0.393*n2))
)
r.Dermatitis <- NormModVar$new(
  'Baseline dermatitis risk', '/catheter', mu=0.0026, sigma=0.00026
)
rr.Dermatitis <- ExprModVar$new(
  "Tegaderm LSI HR", 
  "HR", 
  rlang::quo(exp(1.482-0.490*n3))
)

# cost variables
c.CRBSI <- GammaModVar$new(
  'CRBSI cost', 'GBP', shape=198.0, scale=50
)
c.Dermatitis <- GammaModVar$new(
  'Dermatitis cost', 'GBP', shape=30, scale=5
)
c.LSI <- GammaModVar$new(
  'LSI cost', 'GBP', shape=50, scale=5
)
n.catheters <- NormModVar$new(
  'No. catheters', 'catheters', mu=3, sigma=0.3 
)
c.Tegaderm <- ExprModVar$new(
  "Tegaderm CHG cost", "GBP", rlang::quo(6.21*n.catheters)
)
c.Standard <- ExprModVar$new(
  "Standard dressing cost", "GBP", rlang::quo(1.34*n.catheters)
)
n.cathdays <- NormModVar$new(
  'No. days with catheter', 'days', mu=10, sigma=2
)  
```

# Model variable expressions
Variables in the model may be included in the decision tree via mathematical 
expressions, which involve model variables and are themselves
model variables. Forms of expression involving R's numerical functions and
multiple model variables are supported, provided they conform to R syntax.
The following code creates the model variable expressions to be used as values
in the decision tree edges. For probabilities, the convention $q = 1-p$ is 
used to ensure that the sum of probabilities leaving each chance node sums
to one.

```{r expressions, echo=TRUE}
# probabilities
p.Dermatitis.S <- ExprModVar$new(
  'P(dermatitis|standard dressing)', 'P', 
  rlang::quo(r.Dermatitis*n.catheters)
)
q.Dermatitis.S <- ExprModVar$new(
  'Q(dermatitis|standard dressing)', '1-P', 
  rlang::quo(1-p.Dermatitis.S)
)
p.Dermatitis.T <- ExprModVar$new(
  'P(dermatitis|Tegaderm)', 'P', 
  rlang::quo(r.Dermatitis*rr.Dermatitis*n.catheters)
)
q.Dermatitis.T <- ExprModVar$new(
  'Q(dermatitis|Tegaderm)', '1-P', 
  rlang::quo(1-p.Dermatitis.T)
)

p.LSI.S <- NormModVar$new(
  'P(LSI|Standard)', '/patient', mu=0.1, sigma=0.01 
)
q.LSI.S <- ExprModVar$new(
  'Q(LSI|Standard)', '1-P', rlang::quo(1-p.LSI.S) 
)
p.LSI.T <- ExprModVar$new(
  'P(LSI|Tegaderm)', 'P', rlang::quo(1-(1-p.LSI.S)^hr.LSI)
)
q.LSI.T <- ExprModVar$new(
  'Q(LSI|Tegaderm)', '1-P', rlang::quo(1-p.LSI.T)
)

p.CRBSI.S <- ExprModVar$new(
  'P(CRBSI|standard dressing)', 'P',  rlang::quo(r.CRBSI*n.cathdays/1000)
)
q.CRBSI.S <- ExprModVar$new(
  'Q(CRBSI|standard dressing)', '1-P',  rlang::quo(1-p.CRBSI.S)
)
p.CRBSI.T <- ExprModVar$new(
  'P(CRBSI|Tegaderm)', 'P', rlang::quo(1-(1-r.CRBSI*n.cathdays/1000)^hr.CRBSI)
)
q.CRBSI.T <- ExprModVar$new(
  'Q(CRBSI|Tegaderm)', '1-P', rlang::quo(1-p.CRBSI.T)
)
```

# The decision tree
The following code constructs the decision tree based on figure 2
of Jenks *et al* [-@jenks2016]. In the formulation used by `rdecision`, 
the decision tree is constructed from sets of decision, chance and 
leaf nodes and from edges (actions and reactions).
Leaf nodes are synonymous with
pathways in Briggs' terminology [-@briggs2006]. The time horizon is
not stated explicitly in the model, and is assumed to be 7 days. It was implied
that the time horizon was ICU stay plus some follow-up, and the costs reflect
those incurred in that period, so the assumption of 7 days does not affect
the `rdecision` implementation of the model.

The tree is somewhat more complex than figure 2 of Jenks *et al* because it
allows for patients to have more than one adverse event (AE) during their stay
(whereas figure 2 implies that only one event per patient is possible). The
rates of AE were estimated independently, and allow for multiple events.

```{r tree, echo=T}
# create decision tree
th <- as.difftime(7, units="days")
# standard dressing
t01 <- LeafNode$new("t01", interval=th)
t02 <- LeafNode$new("t02", interval=th)
c01 <- ChanceNode$new()
e01 <- Reaction$new(c01,t01,p=p.Dermatitis.S,cost=c.Dermatitis,
                    label="Dermatitis")
e02 <- Reaction$new(c01,t02,p=q.Dermatitis.S,cost=0,
                    label="No dermatitis")
#
t03 <- LeafNode$new("t03", interval=th)
t04 <- LeafNode$new("t04", interval=th)
c02 <- ChanceNode$new()
e03 <- Reaction$new(c02,t03,p=p.Dermatitis.S,cost=c.Dermatitis,
                    label="Dermatitis")
e04 <- Reaction$new(c02,t04,p=q.Dermatitis.S,cost=0,
                    label="No dermatitis")
#
c03 <- ChanceNode$new()
e05 <- Reaction$new(c03,c01,p=p.LSI.S,cost=c.LSI,label="LSI")
e06 <- Reaction$new(c03,c02,p=q.LSI.S,cost=0,label="No LSI")
#
t11 <- LeafNode$new("t11", interval=th)
t12 <- LeafNode$new("t12", interval=th)
c11 <- ChanceNode$new()
e11 <- Reaction$new(c11,t11,p=p.Dermatitis.S,cost=c.Dermatitis,
                    label="Dermatitis")
e12 <- Reaction$new(c11,t12,p=q.Dermatitis.S,cost=0,label="No Dermatitis")
#
t13 <- LeafNode$new("t13", interval=th)
t14 <- LeafNode$new("t14", interval=th)
c12 <- ChanceNode$new()
e13 <- Reaction$new(c12,t13,p=p.Dermatitis.S,cost=c.Dermatitis, 
                    label="Dermatitis")
e14 <- Reaction$new(c12,t14,p=q.Dermatitis.S,cost=0,label="No dermatitis")
#
c13 <- ChanceNode$new()
e15 <- Reaction$new(c13,c11,p=p.LSI.S,cost=c.LSI,label="LSI")
e16 <- Reaction$new(c13,c12,p=q.LSI.S,cost=0,label="No LSI")
#
c23 <- ChanceNode$new()
e21 <- Reaction$new(c23,c03,p=p.CRBSI.S,cost=c.CRBSI,label="CRBSI")
e22 <- Reaction$new(c23,c13,p=q.CRBSI.S,cost=0,label="No CRBSI")
#
# Tegaderm branch  
t31 <- LeafNode$new("t31", interval=th)
t32 <- LeafNode$new("t32", interval=th)
c31 <- ChanceNode$new()
e31 <- Reaction$new(c31,t31,p=p.Dermatitis.T,cost=c.Dermatitis,
                    label="Dermatitis")
e32 <- Reaction$new(c31,t32,p=q.Dermatitis.T,cost=0,label="no dermatitis")
#
t33 <- LeafNode$new("t33", interval=th)
t34 <- LeafNode$new("t34", interval=th)
c32 <- ChanceNode$new()
e33 <- Reaction$new(c32,t33,p=p.Dermatitis.T,cost=c.Dermatitis,
                    label="Dermatitis")
e34 <- Reaction$new(c32,t34,p=q.Dermatitis.T,cost=0,label="No dermatitis")
#
c33 <- ChanceNode$new()
e35 <- Reaction$new(c33,c31,p=p.LSI.T,cost=c.LSI,label="LSI")
e36 <- Reaction$new(c33,c32,p=q.LSI.T,cost=0,label="No LSI")
#
t41 <- LeafNode$new("t41", interval=th)
t42 <- LeafNode$new("t42", interval=th)
c41 <- ChanceNode$new()
e41 <- Reaction$new(c41,t41,p=p.Dermatitis.T,cost=c.Dermatitis,
                    label="Dermatitis")
e42 <- Reaction$new(c41,t42,p=q.Dermatitis.T,cost=0,label="No dermatitis")
#
t43 <- LeafNode$new("t43", interval=th)
t44 <- LeafNode$new("t44", interval=th)
c42 <- ChanceNode$new()
e43 <- Reaction$new(c42,t43,p=p.Dermatitis.T,cost=c.Dermatitis,
                    label="Dermatitis")
e44 <- Reaction$new(c42,t44,p=q.Dermatitis.T,cost=0,label="No dermatitis")
#
c43 <- ChanceNode$new()
e45 <- Reaction$new(c43,c41,p=p.LSI.T,cost=c.LSI,label="LSI")
e46 <- Reaction$new(c43,c42,p=q.LSI.T,cost=0,label="No LSI")
#
c53 <- ChanceNode$new()
e51 <- Reaction$new(c53,c43,p=p.CRBSI.T,cost=c.CRBSI,label="CRBSI")
e52 <- Reaction$new(c53,c33,p=q.CRBSI.T,cost=0,label="no CRBSI")
#  
# decision node
d1 <- DecisionNode$new("d1")
e9 <- Action$new(d1,c23,label="Standard",cost=c.Standard)
e10 <- Action$new(d1,c53,label="Tegaderm",cost=c.Tegaderm)
#
# create decision tree
V <- list(d1,c01,c02,c03,c11,c12,c13,c23,c31,c32,c33,c41,c42,c43,c53,
          t01,t02,t03,t04,t11,t12,t13,t14,t31,t32,t33,t34,t41,t42,t43,t44)
E <- list(e01,e02,e03,e04,e05,e06,e11,e12,e13,e14,e15,e16,e21,e22,
          e31,e32,e33,e34,e35,e36,e41,e42,e43,e44,e45,e46,e51,e52,e9,e10)
DT <- DecisionTree$new(V,E)
```

```{r draw, echo=FALSE, results="hide", fig.keep="last", fig.align="center"}
DT$draw(border=TRUE)
```

In the manufacturer's model, the uncertainties in the probabilities associated
with the polytomous chance nodes were modelled as independent variables. This is
not recommended because there is a chance that a particular run of the PSA will
yield probabilities that are outside the range [0,1]. Representing the 
uncertain probabilities with draws from a Dirichlet distribution is preferred. 

# Summary of the model
The model variables associated with actions, reactions and leaf nodes
can be tabulated using the method `modvar_table`. This returns a data
frame describing each variable, its description, units and uncertainty
distribution. Variables inheriting from type `ModVar` will be included in the
tabulation; regular numeric values will not be listed. In the Tegaderm 
model, the complete structure
is as follows:

```{r modelinputs-structure, echo=FALSE}
local({
  DF <- DT$modvar_table()
  keep <- c("Description", "Distribution")
  pander::pander(DF[,keep], row.names=F, digits=3, justify="left")
})
```

# Point estimates and distributions of model variables
The point estimates, units and distributional properties are obtained from the
same call, in the remaining columns. 

```{r modelinputs-pe, echo=FALSE}
local({
  DF <- DT$modvar_table()
  keep <- c("Description", "Units", "Mean", "Q2.5", "Q97.5")
  pander::pander(DF[,keep], row.names=F, digits=3, justify="left")
})
```

# Running the model: base case
The following code runs a single model scenario, using the `evaluate`
method of a decision node to evaluate each pathway from the decision node,
shown in the table. This model did not consider utility, and the columns 
associated with utility are removed. 

```{r basecase, echo=TRUE}
RES <- DT$evaluate()
```

```{r basecase_table, echo=FALSE}
local({
  keep <- c('Run', 'd1', 'Cost')
  pander::pander(RES[,keep], round=2, justify="llr")
})
```

# Running the model: probabilistic sensitivity analysis
Probabilistic sensitivity analysis is supported through the use of sampling 
model variables. The same call, with extra parameters is used to run the PSA:

```{r PSA, echo=TRUE}
N <- 1000
PSA <- DT$evaluate(setvars="random", N=N)
```

The first few runs of PSA are as follows (after reshaping the table to give
one row per simulation, rather than one row per run, per strategy):

```{r PSA_table, echo=FALSE}
local({
  RES <<- reshape(PSA, idvar='Run', timevar='d1', direction='wide')
  RES$Difference <<- RES$Cost.Tegaderm - RES$Cost.Standard
  keep<- c('Run', 'Cost.Tegaderm', 'Cost.Standard', 'Difference')
  pander::pander(head(RES[,keep], n=10), round=2, row.names=F, justify="lrrr")
})
```

From PSA (`r N` runs), the mean cost of treatment with Tegaderm 
was `r round(mean(RES$Cost.Tegaderm),2)`,
the mean cost of treatment with standard dressings was 
`r round(mean(RES$Cost.Standard),2)`
and the mean cost saving was `r round(mean(RES$Difference),2)`. The 95% 
confidence interval for cost saving was 
`r round(quantile(RES$Difference, probs=c(0.025)),2)` to 
`r round(quantile(RES$Difference, probs=c(0.975)),2)`; the standard deviation
of the cost saving was `r round(sd(RES$Difference),2)`.
Overall, `r round(100*sum(RES$Difference<0)/nrow(RES),2)`% of runs found that
Tegaderm was cost saving. These results replicate those reported by the
manufacturer (saving of 77.76, 98.5% cases cost saving; mean cost of standard 
dressing 176.89, mean cost of Tegaderm 99.63).

```{r echo=F}
rm(RES)
```



# References
