---
title: "Shale gas"
subtitle: "A decision tree example"
author: "Andrew J. Sims"
date: "July 2020"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Shale gas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "references.bib"
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=F}
library(rdecision)
```

# Introduction
Kamiński *et al* [-@kaminski:2018a, Fig 7] provide an example of a decision tree
with multiple decision nodes, including some that are descendants of another
decision node. This vignette illustrates how `rdecision` can be used to model
a complex decision tree, using the example from Figure 7 of Kamiński 
*et al* [-@kaminski:2018a]. 

# The problem
Kamiński *et al* [-@kaminski:2018a] state the problem as follows:

> Consider an investor owning a plot of land, possibly (*a priori* probability
> amounting to 70%) hiding shale gas layers. The plot can be sold immediately
> (800, all prices in $'000). The investor can build a gas extraction unit for 
> a cost of 300. If gas is found, the profit will amount to 2,500 (if not there
> will be no profit, and no possibility of selling the land). Geological tests
> can be performed for a cost of 50, and will produce either a positive or
> negative signal. The sensitivity amounts to 90% and the specificity amounts
> to 70%. The installation can be built after the test or the land may be sold
> for 1000 (600) after a positive (negative) test result.

# Creating the model
The model, comprising three decision nodes, four chance nodes, nine leaf
nodes and 15 edges, is constructed as follows. Costs, benefits and 
probabilities are associated with each edge, which must be an Action
or a Reaction object.

```{r create-model, echo=TRUE}
  # nodes
  d1 <- DecisionNode$new("d1")
  d2 <- DecisionNode$new("d2")
  d3 <- DecisionNode$new("d3")
  c1 <- ChanceNode$new("c1")
  c2 <- ChanceNode$new("c2")
  c3 <- ChanceNode$new("c3")
  c4 <- ChanceNode$new("c4")
  t1 <- LeafNode$new("t1")
  t2 <- LeafNode$new("t2")
  t3 <- LeafNode$new("t3")
  t4 <- LeafNode$new("t4")
  t5 <- LeafNode$new("t5")
  t6 <- LeafNode$new("t6")
  t7 <- LeafNode$new("t7")
  t8 <- LeafNode$new("t8")
  t9 <- LeafNode$new("t9")
  # probabilities
  p.sens <- 0.9
  p.spec <- 0.7
  p.gas <- 0.7
  p.nogas <- 1-p.gas
  p.ptest <- p.sens*p.gas + (1-p.spec)*p.nogas
  p.ntest <- (1-p.sens)*p.gas + p.spec*p.nogas
  p.gas.ptest <- p.sens*p.gas / p.ptest
  p.gas.ntest <- (1-p.sens)*p.gas / p.ntest
  # edges
  E <- list(
    Action$new(d1,t1,"sell",benefit=800),
    Action$new(d1,c1,"dig",cost=300),
    Reaction$new(c1,t2,p=p.gas,benefit=2500,label="gas"),
    Reaction$new(c1,t3,p=p.nogas,label="no gas"),
    Action$new(d1,c2,"test",cost=50),
    Reaction$new(c2,d2,p=p.ntest,label="negative"),
    Action$new(d2,t4,"sell",benefit=600),
    Action$new(d2,c3,"dig",cost=300),
    Reaction$new(c3,t5,p=p.gas.ntest,benefit=2500,label="gas"),
    Reaction$new(c3,t6,p=(1-p.gas.ntest),label="no gas"),
    Reaction$new(c2,d3,p=p.ptest,label="positive"),
    Action$new(d3,t7,"sell",benefit=1000),
    Action$new(d3,c4,"dig",cost=300),
    Reaction$new(c4,t8,p=p.gas.ptest,benefit=2500,label="gas"),
    Reaction$new(c4,t9,p=(1-p.gas.ptest),label="no gas")
  )
  # tree
  V <- list(d1,d2,d3, c1,c2,c3,c4, t1,t2,t3,t4,t5,t6,t7,t8,t9)
  DT<-DecisionTree$new(V,E)
  
```

# Evaluating the strategies
There are a total of 12 possible strategies (3 choices from node `d1`
$\times$ 2 choices at node `d2` $\times$ 2 choices at node `d3`). But 
some of these are not unique. For example if the choice at node `d1` is
"sell", the choices at nodes `d2` and `d3` (4 possible combinations) are
unimportant; all four such strategies are identical. In identifying
strategies, `rdecision` discards non-unique strategies, and uses the 
labels of an arbitrarily chosen strategy within each identical
group as the identifier for the strategy; i.e. "sell/sell/sell" also
represents the identical "sell/sell/dig", "sell/dig/sell" and 
"sell/dig/dig".

Similarly, four strategies with choice "dig" at node `d1` are
identical. Removing the six repeated strategies leaves a total
of six possible strategies to evaluate.

Unique strategies are identified by `rdecision` automatically.
Method `evaluate` calculates the expected cost, benefit and
utility of each traversable path for each strategy, and aggregates
by strategy. The results for the gas problem are computed
as follows. Payoff is defined as benefit minus cost. 

```{r evaluate, echo=TRUE}
# find optimal strategies
RES <- DT$evaluate()
RES$Payoff <- RES$Benefit-RES$Cost
```

This gives the following payoff for each unique strategy:

```{r results, echo=FALSE, results='asis'}
RES$Run <- NULL
RES$Probability <- NULL
RES$Utility <- NULL
knitr::kable(RES, row.names=FALSE)
```

```{r poptimal, echo=FALSE}
imax <- which.max(RES$Payoff)
popt <- paste(RES$d1[imax], RES$d2[imax], RES$d3[imax], sep="/")
```
The optimal strategy is `r popt`, *i.e.* test, sell if negative and
dig otherwise. The expected payoff from this strategy is
`r RES$Payoff[imax]`.

# References


