## Model results

### Base case
```{r basecase-summary, echo=FALSE}
PE.saving <- 0
local({
  SUM <- d$evaluateChoices(expected=T)
  PE.saving <<- SUM$Cost[SUM$Choice=='Tegaderm'] - 
                SUM$Cost[SUM$Choice=='Standard']
})
```
The total cost for each choice can be calculated from the table above, or by calling
`evaluateChoices`, giving a point estimate of the saving of `r round(-1*PE.saving,2)`
GBP. This is close to the sponsor's point 
estimate of cost saving estimated from their probabilistic sensitivity analysis,
77.26 GBP, reported in Jenks *et al* [-@jenks:2016a].

### Probabilistic senstivity analysis
When they are created, each `ModVar` returns its expected value when its
method `value()` is called. Calling the method `sample()` of a model variable causes
it to sample from its uncertainty distribution, and return the sampled value when
method `value()` is next called. The same sampled value will be returned until
`sample()` is called again. Calling `sample(expected=T)` causes `value()` to 
return the expected value of the variable. 

Probabilistic sensitivity analysis
is supported through the use of sampling model variables. In practice, because the
model variables enter the model via nodes, the methods `evaluatePathways` and
`evaluateChoices` provided by decision nodes provide a convenient interface for
sampling from model variables. These methods, called with `expected=FALSE` cause
each model variable associated with the decision node and its descendants to be
sampled. For further convenience the method `evaluateChoices` permits replicates
to be run (parameter `N`), making PSA straightforward. The first few runs of
PSA are as follows:

```{r PSA, echo=FALSE}
local({
  PSA <- d$evaluateChoices(expected=F, uncorrelate=F, N=1000)
  RES <<- reshape(PSA, idvar='Run', timevar='Choice', direction='wide')
  RES$Difference <<- RES$Cost.Tegaderm - RES$Cost.Standard
  keep<- c('Run', 'Cost.Tegaderm', 'Cost.Standard', 'Difference')
  knitr::kable(head(RES[,keep], n=10), digits=2, row.names=F)
})
```

From PSA (1000 runs), the mean cost of treatment with Tegaderm 
was `r round(mean(RES$Cost.Tegaderm),2)`,
the mean cost of treatment with standard dressings was `r round(mean(RES$Cost.Standard),2)`
and the mean cost saving was `r round(mean(RES$Difference),2)`. The 95% confidence 
interval for cost saving was `r round(quantile(RES$Difference, probs=c(0.025)),2)` to 
`r round(quantile(RES$Difference, probs=c(0.975)),2)`; the standard deviation of the
cost saving was `r round(sd(RES$Difference),2)`.
Overall, `r round(100*sum(RES$Difference<0)/nrow(RES),2)`% of runs found that Tegaderm
was cost saving. These results replicate those reported by the manufucturer (saving
of 77.26, 98.5% cases cost saving).

```{r echo=F}
rm(RES)
```

# An alternative, tree-free approach
It is possible to solve the decision problem without first constructing a tree, by
combining model variables directly. This is, in essence, the approach taken in
many decision tree models constructed in Excel. 

## Components of cost
Each cost component is defined as an expression involving two or more
of the 13 model inputs. In contrast to the tree approach, which computed
the cost of traversing each pathway, this approach allows the costs of the
technology and its comparator to be constructed as sub-totals.
```{r treefree, echo=TRUE}
# component costs, standard dressing
CHG.S <- ExprModVar$new(
  "Cost std dressing", "GBP", quote(n.dressings*c.Standard)
)
CRBSI.S <- ExprModVar$new(
  "Cost CRBSI, std dressing", "GBP", quote(c.CRBSI*r.CRBSI*n.cathdays/1000)
)
LSI.S <- ExprModVar$new(
  "Cost LSI, std dressing", "GBP", quote(c.LSI*r.LSI)
)
Dermatitis.S <- ExprModVar$new(
  "Cost dermatitis, std dressing", "GBP", quote(r.Dermatitis*c.Dermatitis*n.dressings)
)

# component costs, Tegaderm
CHG.T <- ExprModVar$new(
  "Cost Tegaderm", "GBP", quote(n.dressings*c.Tegaderm)
)
CRBSI.T <- ExprModVar$new(
  "Cost CRBSI, Tegaderm", "GBP", quote(c.CRBSI*r.CRBSI*hr.CRBSI*n.cathdays/1000)
)
LSI.T <- ExprModVar$new(
  "Cost LSI, Tegaderm", "GBP", quote(c.LSI*r.LSI*hr.LSI)
)
Dermatitis.T <- ExprModVar$new(
  "Cost dermatitis, Tegaderm", "GBP", 
   quote(r.Dermatitis*c.Dermatitis*rr.Dermatitis*n.dressings)
)

# per-patient costs
total.T <- ExprModVar$new(
  'Treatment cost, Tegaderm', 'GBP', quote(CHG.T+CRBSI.T+LSI.T+Dermatitis.T)
)
total.S <- ExprModVar$new(
  'Treatment cost, Standard', 'GBP', quote(CHG.S+CRBSI.S+LSI.S+Dermatitis.S)
)
c.diff <- ExprModVar$new(
  'Cost difference', 'GBP', quote(total.T-total.S)
)
```

## Base case
The components of cost can be extracted by tabulating the model variables
in the cost difference model variable, `c.diff` via method `tabulate`. In this
case, only the rows containing component costs are displayed.
```{r treefree-base, echo=FALSE}
local({
  MV <- c.diff$tabulate(include.operands=TRUE)
  vars <- c('CHG.S', 'CRBSI.S', 'LSI.S', 'Dermatitis.S',
            'CHG.T', 'CRBSI.T', 'LSI.T', 'Dermatitis.T',
            'total.S', 'total.T', 'c.diff')
  keep <- c('Description', 'Units', 'Mean', 'SD', 'Q2.5', 'Q97.5', 'Qhat')
  knitr::kable(MV[MV$Label %in% vars, keep], row.names=F, 
               format.args=list(scientific=F), digits=2
               )
})
```

The point estimate of saving is obtained directly from the expectation
of the cost difference variable, `r round(-1*c.diff$getMean(),2)`. This
is identical to the value obtained from the full tree. The table also
indicates, by the quantiles of the cost difference, the approximate
confidence interval of the saving.

## Probabilistic sensitivity analysis
```{r echo=FALSE}
tf.psa <- c.diff$r(1000)
```
Each model variable provides a method, `r`, to make random draws from its
uncertainty distribution. The PSA for the cost difference can therefore be 
achieved by a single call to this method. From this, the mean cost saving
was `r round(mean(tf.psa),2)`, the 
95% confidence interval of the saving was `r round(quantile(tf.psa,0.025),2)` 
to `r round(quantile(tf.psa,0.975),2)` 
and `r round(100*sum(tf.psa<0)/length(tf.psa),2)`% of runs generated a cost saving.
Within fluctuation error this is consistent with the manufacturer's reported
saving of 77.26 with 98.5% of runs being cost saving.

# A note on correlation of model variables
Of the 13 input variables defined in the tree model, eight appear in both the
main branches of the decision tree. If, in PSA, common variables are sampled
from their uncertainty distributions once per run, rather being sampled for 
each choice within a run, the costs of each arm will be correlated.

```{r CORR, echo=FALSE}
local({
  PSA <- d$evaluateChoices(expected=F, uncorrelate=F, N=1000)
  RES.C <<- reshape(PSA, idvar='Run', timevar='Choice', direction='wide')
  RES.C$Difference <<- RES.C$Cost.Tegaderm - RES.C$Cost.Standard
  R.C <<- cor(RES.C$Cost.Tegaderm, RES.C$Cost.Standard)

  PSA <- d$evaluateChoices(expected=F, uncorrelate=T, N=1000)
  RES.U <<- reshape(PSA, idvar='Run', timevar='Choice', direction='wide')
  RES.U$Difference <<- RES.U$Cost.Tegaderm - RES.U$Cost.Standard
  R.U <<- cor(RES.U$Cost.Tegaderm, RES.U$Cost.Standard)
  
  sd.T <<- sd(RES.C$Cost.Tegaderm)
  sd.S <<- sd(RES.C$Cost.Standard)
  cov.TS <<- cov(RES.C$Cost.Tegaderm, RES.C$Cost.Standard)  
})
```

In the original Excel model submitted by the manufacturer, the model variables
were sampled once per run. The tree model constructed using `rdecision` described
in this vignette made the same assumption, and replicated the manufacturer's
results. The correlation coefficient between 1000 samples from the cost of 
treatment with Tegaderm and the cost with standard care was `r round(R.C, 3)`. This
reflects the relatively high degree of correlation between the two choice
arms of the model, because of the eight shared model variables.

However, `rdecision` includes the facility to resample the model
variables for each choice, within each run, by setting parameter
`uncorrelate=TRUE` in method `evaluateChoices` of a decision node.
With this option, the equivalent correlation coefficient is `r round(R.U, 3)`;
*i.e.* they are uncorrelated within statistical fluctuation. This does not
affect the point estimate, but does alter the distribution of the
cost difference, which has standard deviation `r round(sd(RES.U$Difference),2)`
and 95% confidence interval `r round(quantile(RES.U$Difference, probs=c(0.025)),2)` 
to `r round(quantile(RES.U$Difference, probs=c(0.975)),2)`, 
with `r round(100*sum(RES.U<0)/nrow(RES.U),1)`% of simulations being
cost saving.
The confidence interval includes zero, but the percentage of
cost saving cases exceeds 95%; with correlation removed, the evidence
for a cost saving is therefore marginal, given the input parameters.

Finally, the variance of the cost of treatment with Tegaderm 
is `r round(sd.T,2)`$^2$, the variance of the cost of treatment with
standard care is `r round(sd.S,2)`$^2$ and the covariance with
correlated model variables is `r round(cov.TS,2)`. Via the relation
$Var(X-Y) = Var(X) + Var(Y) - 2Cov(X,Y)$, the variance of the
cost difference with correlated model variables 
is `r round(sqrt(sd.T^2 + sd.S^2 - 2*cov.TS),2)`$^2$ and with the
uncorrelated model variables 
is `r round(sqrt(sd.T^2 + sd.S^2),2)`$^2$, close to the observed values.

```{r CORREND, echo=FALSE}
rm(RES.C, RES.U, R.C, R.U, sd.T, sd.S, cov.TS)
```

