---
title: "Elementary Markov Model (Chancellor 1997)"
subtitle: "Monotherapy versus combination therapy for HIV"
author: "Andrew J. Sims"
date: "May 2021"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Elementary Markov Model (Chancellor 1997)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/national-institute-of-health-research.csl"
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.keep = "last",
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=F}
library(rdecision)
library(pander)
library(png)
```

# Introduction
This vignette is an example of an elementary cohort Markov model using 
the `rdecision` package. It is based on the example given by 
Briggs *et al* [-@briggs2006] (Exercise 2.5) which itself is based on a Markov
model described by Chancellor *et al* [@chancellor1997]. The model compares 
a combination therapy of Lamivudine/Zidovudine versus Zidovudine monotherapy 
in people with HIV infection.

# Creating the model
The variables used in the model are all numerical constants, and are defined
as follows. The original model was based on annual transition probabilities;
these are converted to instantaneous hazard rates in units of events/year.
```{r model-variables, echo=TRUE}
# transition rates calculated from annual transition probabilities
trAB <- -log(1-0.202)/1 
trAC <- -log(1-0.067)/1
trAD <- -log(1-0.010)/1
trBC <- -log(1-0.407)/1
trBD <- -log(1-0.012)/1
trCD <- -log(1-0.250)/1
# Costs
dmca <- 1701 # direct medical costs associated with state A
dmcb <- 1774 # direct medical costs associated with state B
dmcc <- 6948 # direct medical costs associated with state C
ccca <- 1055 # Community care costs associated with state A
cccb <- 1278 # Community care costs associated with state B
cccc <- 2059 # Community care costs associated with state C
# Drug costs
cAZT <- 2278 # zidovudine drug cost
cLam <- 2086 # lamivudine drug cost
# Other parameters
RR <- 0.509 # treatment effect
cDR <- 6 # annual discount rate, costs (%)
oDR <- 0 # annual discount rate, benefits (%)
```

The model is constructed by forming a graph, with each state as a 
node and each transition as an edge. Nodes (of class `MarkovState`) and edges
(class `MarkovTransition`) have various properties whose values reflect the
variables of the model (costs, rates etc.). The rate for one of the outgoing
transitions from each non-absorbing state is set to NULL to allow the sum
of probabilities leaving each state, per cycle, to be adjusted to 1 within the
package. The usual case, as here, is to set the self-loop rates to NULL and the 
package will compute the probability of remaining in a state as one minus the 
probability of leaving the state. Because the model is intended to evaluate 
survival, the utility of states A, B and C are set to 1 (by default) and
state D to zero. Thus the incremental quality adjusted life years gained per
cycle is equivalent to the survival function. 
```{r model, echo=TRUE}
# create Markov states for monotherapy (zidovudine only)
sA <- MarkovState$new("A", cost=dmca+ccca+cAZT)
sB <- MarkovState$new("B", cost=dmcb+cccb+cAZT)
sC <- MarkovState$new("C", cost=dmcc+cccc+cAZT)
sD <- MarkovState$new("D", cost=0, utility=0)
# create transitions
tAA <- MarkovTransition$new(sA, sA, r=NULL)
tAB <- MarkovTransition$new(sA, sB, r=trAB)
tAC <- MarkovTransition$new(sA, sC, r=trAC)
tAD <- MarkovTransition$new(sA, sD, r=trAD)
tBB <- MarkovTransition$new(sB, sB, r=NULL)
tBC <- MarkovTransition$new(sB, sC, r=trBC)
tBD <- MarkovTransition$new(sB, sD, r=trBD)
tCC <- MarkovTransition$new(sC, sC, r=NULL)
tCD <- MarkovTransition$new(sC, sD, r=trCD)
tDD <- MarkovTransition$new(sD, sD, r=NULL)
# construct the model
mhiv <- CohortMarkovModel$new(
  V = list(sA, sB, sC, sD),
  E = list(tAA, tAB, tAC, tAD, tBB, tBC, tBD, tCC, tCD, tDD),
  hcc = FALSE,
  discount.cost = cDR/100,
  discount.utility = oDR/100
)
```

# Checking the model

## Diagram
A representation of the model in DOT format ([Graphviz](https://graphviz.org))
can be created using the `as_DOT` function of `CohortMarkovModel`. The function
returns a character vector which can be saved in a file (`.gv` extension) for
visualization with the `dot` tool of Graphviz, or plotted directly in R via
the `DiagrammeR` package. The Markov model for monotherapy is shown in Figure 1.

```{r caption, echo=FALSE}
f1c  <- paste("Markov model for comparison of HIV therapy.", 
              "A: 200 < cd4 < 500,", "B: cd4 < 200,", "C: AIDS,","D: Death.")
```
```{r draw,results="hide",fig.cap=f1c,fig.asp=0.21}
DOT <- mhiv$as_DOT()
writeLines(DOT, con="mono.gv")
system2(command="dot", args=c("-Tpng","-o","mono.png","mono.gv"))
img <- png::readPNG(source="mono.png", native=TRUE)
grid::grid.newpage()
grid::grid.raster(image=img)
```

## Model states
The states in the model can be tabulated with the function `tabulate_states`.
For the monotherapy model, the states are tabulated below. The cost of each
state includes the annual cost of AZT (Zidovudine). 
```{r echo=FALSE}
DF <- mhiv$tabulate_states()
pander::pander(DF[,c("Name", "Cost")], justify="lr")
rm(DF)
```

## Per-cycle transition probabilities
The per-cycle transition probabilities, which are the cells of the Markov
transition matrix, can be extracted from the model via the function
`transition_probability`. For the monotherapy model, the transition matrix is
shown below. This is consistent with the Table 1 of 
Chancellor *et al* [-@chancellor1997].
```{r echo=FALSE}
TM <- mhiv$transition_probability()
pander::pander(TM, emphasize.rownames=FALSE, justify="lcccc")
rm(TM)
```

# Running the model

Model function `cycle` applies one cycle of a Markov model to a defined 
starting population in each state. It returns a table with one row per state, 
and each row containing several columns, including the population at the end of
the state and the cost of occupancy of states, normalized by the number of 
patients in the cohort, with discounting applied.

Multiple cycles are run by feeding the state populations at the end of
one cycle into the next. Function `cycles` does this and returns a data frame 
with one row per cycle, and each row containing the state populations and the
aggregated cost of occupancy for all states, with discounting applied. This is
done below for the first 20 cycles of the model for monotherapy, without half
cycle correction, with discount. In addition, the proportion of patients alive 
at each cycle (the Markov trace) is added to the table. The populations and
discounted costs are consistent with Briggs *et al*, Table 2.3 [-@briggs2006], 
and the QALY column is consistent with Table 2.4 (without half cycle
correction). No discount was applied to the utilities.
```{r monocycle, echo=TRUE}
# create starting populations
N <- 1000
populations <- c(A = N, B = 0, C = 0, D = 0)
mhiv$set_populations(populations)
# run 20 cycles
MT.mono <- mhiv$cycles(ncycles=20)
```

```{r print_monocycle}
keep <- c("Years", "A", "B", "C", "D", "Cost", "QALY")
pander::pander(MT.mono[,keep], row.names=FALSE, justify="rrrrrrr", 
               round=c(2,0,0,0,0,0,3))
```

# Model results

## Monotherapy
```{r mon_results, echo=FALSE}
el.mono <- sum(MT.mono$QALY)
cost.mono <- sum(MT.mono$Cost)
```
The estimated life years is approximated by summing the proportions of patients
left alive at each cycle (Briggs *et al*[@briggs2006], Exercise 2.5). This is 
an approximation because it ignores the population who remain alive after 
21 years, and assumes all deaths occurred at the start of each cycle. For
monotherapy the expected life gained is `r round(el.mono,3)` years at a cost
of `r format(cost.mono,digits=2,scientific=FALSE)` GBP.

## Combination therapy
For combination therapy, the same model structure was used, and in `rdecision`,
the same nodes, edges and model used for the monotherapy can be re-used for the 
combination therapy model by setting costs and transition rates to amended 
values.
```{r combo, echo=TRUE}
# set occupancy costs for combination therapy (zidovudine and lamivudine)
sA$set_cost(dmca+ccca+cAZT+cLam)
sB$set_cost(dmcb+cccb+cAZT+cLam)
sC$set_cost(dmcc+cccc+cAZT+cLam)
# modify transition rates by the treatment effect
tAB$set_rate(trAB*RR)
tAC$set_rate(trAC*RR)
tAD$set_rate(trAD*RR)
tBC$set_rate(trBC*RR)
tBD$set_rate(trBD*RR)
tCD*set_rate(trCD*RR)
```

In this model, lamivudine is given for the first 2 years, with 
the treatment effect assumed to persist for the same period. The
state populations and cycle numbers are retained by the model between 
calls to `cycle` or `cycles` making it easy to change probabilities or
costs during a simulation. The cycle count and elapsed time are reset when
a new population is set. 
```{r combo_run, echo=TRUE}
# run combination therapy model for 2 years
populations <- c('A'=N, 'B'=0, 'C'=0, 'D'=0)
mhiv$set_populations(populations)
# run 2 cycles
MT.comb <- mhiv$cycles(2)
# revise costs and transitions, and run model for next 18 years
#state.comb.A$setAnnualCost(1701+1055+2278)
#state.comb.B$setAnnualCost(1774+1278+2278)
#state.comb.C$setAnnualCost(6948+2059+2278)
#m.comb$setTransitions(I.mono)
#DF.comb <- rbind(DF.comb, m.comb$cycles(nCycles=18))
## calculate the proportion alive at end of each cycle
#DF.comb$Alive <- (DF.comb$A + DF.comb$B + DF.comb$C)/N
```

The cycle history for combination therapy is as follows:
```{r echo=F}
#knitr::kable(DF.comb)
```

## Comparison of treatments
```{r echo=F}
#el.mono <- sum(DF.mono$Alive)
#el.comb <- sum(DF.comb$Alive)
#cost.mono <- sum(DF.mono$Cost)
#cost.comb <- sum(DF.comb$Cost)
#icer <- (cost.comb-cost.mono)/(el.comb-el.mono)
```
The ICER is calculated by running both models and calculating the incremental
cost per life year gained. 

```{r echo=FALSE}
#Over the 20 year time horizon, the expected life years for monotherapy was
#`r round(el.mono,3)` years at a total cost per patient of 
#`r format(cost.mono,digits=2,scientific=F,big.mark=',')` GBP. The expected life years
#with combination therapy
#was `r round(el.comb,3)` at a total cost per patient of 
#`r format(cost.comb,digits=2,scientific=F,big.mark=',')` GBP. 
#The incremental change in life years was `r round(el.comb-el.mono,3)` years
#at an incremental cost of 
#`r format(cost.comb-cost.mono,digits=2,scientific=F,big.mark=',')` GBP,
#giving an ICER of `r round(icer,2)` GBP.
```


# References
