---
title: "Elementary Semi-Markov Model (Chancellor 1997)"
subtitle: "Monotherapy versus combination therapy for HIV"
author: "Andrew J. Sims"
date: "May 2021"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Elementary Semi-Markov Model (Chancellor 1997)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/nature-no-et-al.csl"
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.keep = "last",
  fig.align = "center",
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library("rdecision")
```

# Introduction
This vignette is an example of an elementary semi-Markov model using 
the `rdecision` package. It is based on the example given by 
Briggs *et al* [-@briggs2006] (Exercise 2.5) which itself is based on a
model described by Chancellor *et al* [@chancellor1997]. The model compares 
a combination therapy of Lamivudine/Zidovudine versus Zidovudine monotherapy 
in people with HIV infection.

# Creating the model

## Model variables
The variables used in the model are all numerical constants, and are defined
as follows. 
```{r model-variables, echo=TRUE}
# transition counts
nAA <- 1251L
nAB <- 350L
nAC <- 116L
nAD <- 17L
nBB <- 731L
nBC <- 512L
nBD <- 15L
nCC <- 1312L
nCD <- 437L
# Healthcare system costs
dmca <- 1701.0 # direct medical costs associated with state A
dmcb <- 1774.0 # direct medical costs associated with state B
dmcc <- 6948.0 # direct medical costs associated with state C
ccca <- 1055.0 # Community care costs associated with state A
cccb <- 1278.0 # Community care costs associated with state B
cccc <- 2059.0 # Community care costs associated with state C
# Drug costs
cAZT <- 2278.0 # zidovudine drug cost
cLam <- 2087.0 # lamivudine drug cost
# Treatment effect
RR <- 0.509 
# Discount rates
cDR <- 6.0 # annual discount rate, costs (%)
oDR <- 0.0 # annual discount rate, benefits (%)
```

## Model structure
The model is constructed by forming a graph, with each state as a 
node and each transition as an edge. Nodes (of class `MarkovState`) and edges
(class `Transition`) may have various properties whose values reflect the
variables of the model (costs, rates etc.). Because the model is intended to 
evaluate survival, the utility of states A, B and C are set to 1 (by default) 
and state D to zero. Thus the incremental quality adjusted life years gained per
cycle is equivalent to the survival function. 
```{r model, echo=TRUE}
# create Markov states for monotherapy (zidovudine only)
sAm <- MarkovState$new("A", cost = dmca + ccca + cAZT)
sBm <- MarkovState$new("B", cost = dmcb + cccb + cAZT)
sCm <- MarkovState$new("C", cost = dmcc + cccc + cAZT)
sDm <- MarkovState$new("D", cost = 0.0, utility = 0.0)
# create transitions
tAAm <- Transition$new(sAm, sAm)
tABm <- Transition$new(sAm, sBm)
tACm <- Transition$new(sAm, sCm)
tADm <- Transition$new(sAm, sDm)
tBBm <- Transition$new(sBm, sBm)
tBCm <- Transition$new(sBm, sCm)
tBDm <- Transition$new(sBm, sDm)
tCCm <- Transition$new(sCm, sCm)
tCDm <- Transition$new(sCm, sDm)
tDDm <- Transition$new(sDm, sDm)
# construct the model
m.mono <- SemiMarkovModel$new(
  V = list(sAm, sBm, sCm, sDm),
  E = list(tAAm, tABm, tACm, tADm, tBBm, tBCm, tBDm, tCCm, tCDm, tDDm),
  discount.cost = cDR / 100.0,
  discount.utility = oDR / 100.0
)
```

## Transition rates and probabilities
Briggs *et al* [-@briggs2006] interpreted the observed transition counts 
in 1 year as transition probabilities by dividing counts by the total 
transitions observed from each state. With this assumption, the annual 
(per-cycle) transition probabilities are calculated as follows and applied
to the model via the `set_probabilities` function.
```{r setprobs, echo=TRUE}
nA <- nAA + nAB + nAC + nAD
nB <- nBB + nBC + nBD
nC <- nCC + nCD
Pt <- matrix(
  c(nAA / nA, nAB / nA, nAC / nA, nAD / nA, 
         0.0, nBB / nB, nBC / nB, nBD / nB,
         0.0,      0.0, nCC / nC, nCD / nC,
         0.0,      0.0,      0.0,      1.0),
  nrow = 4L, byrow = TRUE, 
  dimnames = list(
    source = c("A", "B", "C", "D"), target = c("A", "B", "C", "D")
  )
)
m.mono$set_probabilities(Pt)
```

More usually, fully observed transition counts are converted into 
transition rates (rather than probabilities), as described by Welton and 
Ades [-@welton2005]. This is
because counting events and measuring total time at risk includes 
individuals who make more than one transition during the observation time, and
can lead to rates with values which exceed 1. In contrast, the difference
between a census of the number 
of individuals in each state at the start of the interval and another at the
end is directly related to the per-cycle probability. As Miller and 
Homan [-@miller1994], Welton and Ades [-@welton2005], Jones *et al*
[-@jones2017] and others note, conversion between rates and probabilities for 
multi-state Markov models is non-trivial [@jones2017] and care is needed
when modellers calculate probabilities from published rates for use in
`SemiMarkoModel`s.

# Checking the model

## Diagram
A representation of the model in DOT format ([Graphviz](https://graphviz.org))
can be created using the `as_DOT` function of `SemiMarkovModel`. The function
returns a character vector which can be saved in a file (`.gv` extension) for
visualization with the `dot` tool of Graphviz, or plotted directly in R via
the `DiagrammeR` package. The Markov model for monotherapy is shown in Figure 1.

```{r caption, echo = FALSE}
f1c  <- paste("Figure 1. Markov model for comparison of HIV therapy.", 
              "A: 200 < cd4 < 500,", "B: cd4 < 200,", "C: AIDS,","D: Death.")
``` 
```{r draw, fig.cap = f1c, fig.asp = 0.21}
# This vignette is processed by knitr at build time (on the package developer's
# system) and so the dot tool only needs to be available there, not on any
# system used for checking or installing the vignettes, which operate on the
# versions moved to the doc directory at build time. 
pngfile <- "mono.png" 
dotfile <- "mono.gv"
# update the .gv file
DOT <- m.mono$as_DOT()
writeLines(DOT, con = dotfile)
# if graphviz is installed, create the .png file from the .gv file with the dot
# command line tool, otherwise fail gracefully and draw a placeholder image
pdot <- Sys.which("dot")
if (nchar(pdot["dot"]) > 0L) {
  system2(command = pdot["dot"], args = c("-Tpng", "-o", pngfile, dotfile))
} else {
  # create placeholder image
  grDevices::png(pngfile, width = 480L, height = 480L * 0.21)
  grid::grid.newpage()
  grid::grid.move.to(
    x = grid::unit(0.0, "npc"),
    y = grid::unit(0.0, "npc")
  )
  grid::grid.line.to(
    x = grid::unit(1.0, "npc"),
    y = grid::unit(1.0, "npc")
  )
  grid::grid.move.to(
    x = grid::unit(0.0, "npc"),
    y = grid::unit(1.0, "npc")
  )
  grid::grid.line.to(
    x = grid::unit(1.0, "npc"),
    y = grid::unit(0.0, "npc")
  )
  invisible(grDevices::dev.off())
}
knitr::include_graphics(path = pngfile)
```

## Model states
The states in the model can be tabulated with the function `tabulate_states`.
For the monotherapy model, the states are tabulated below. The cost of each
state includes the annual cost of AZT (Zidovudine). 
```{r echo=FALSE}
DF <- m.mono$tabulate_states()
pander::pander(DF[, c("Name", "Cost")], justify = "lr")
rm(DF)
```

## Per-cycle transition probabilities
The per-cycle transition probabilities, which are the cells of the Markov
transition matrix, can be extracted from the model via the function
`transition_probabilities`. For the monotherapy model, the transition matrix is
shown below. This is consistent with the Table 1 of 
Chancellor *et al* [-@chancellor1997].
```{r echo=FALSE}
TM <- m.mono$transition_probabilities()
pander::pander(TM, emphasize.rownames = FALSE, justify = "lcccc")
rm(TM)
```

# Running the model
Model function `cycle` applies one cycle of a Markov model to a defined 
starting population in each state. It returns a table with one row per state, 
and each row containing several columns, including the population at the end of
the state and the cost of occupancy of states, normalized by the number of 
patients in the cohort, with discounting applied.

Multiple cycles are run by feeding the state populations at the end of
one cycle into the next. Function `cycles` does this and returns a data frame 
with one row per cycle, and each row containing the state populations and the
aggregated cost of occupancy for all states, with discounting applied. This is
done below for the first 20 cycles of the model for monotherapy, without half
cycle correction, with discount. In addition, the proportion of patients alive 
at each cycle (the Markov trace) is added to the table. The populations and
discounted costs are consistent with Briggs *et al*, Table 2.3 [-@briggs2006], 
and the QALY column is consistent with Table 2.4 (without half cycle
correction). No discount was applied to the utilities.
```{r monocycle, echo=TRUE}
# create starting populations
N <- 1000L
populations <- c(A = N, B = 0L, C = 0L, D = 0L)
m.mono$reset(populations)
# run 20 cycles
MT.mono <- m.mono$cycles(ncycles = 20L, hcc.pop = FALSE, hcc.cost = FALSE)
```

```{r print_monocycle}
keep <- c("Years", "A", "B", "C", "D", "Cost", "QALY")
pander::pander(MT.mono[, keep], row.names = FALSE, justify = "rrrrrrr", 
               round=c(2L, 0L, 0L, 0L, 0L, 0L, 3L))
```

# Model results

## Monotherapy
```{r mon_results, echo=FALSE}
el.mono <- sum(MT.mono$QALY)
cost.mono <- sum(MT.mono$Cost)
```
The estimated life years is approximated by summing the proportions of patients
left alive at each cycle (Briggs *et al* [@briggs2006], Exercise 2.5). This is 
an approximation because it ignores the population who remain alive after 
21 years, and assumes all deaths occurred at the start of each cycle. For
monotherapy the expected life gained is `r round(el.mono,3)` years at a cost
of `r format(cost.mono,digits=2,scientific=FALSE)` GBP.

## Combination therapy
For combination therapy, a similar model was created, with adjusted costs and 
transition rates. Following Briggs *et al* [@briggs2006] the treatment effect
was applied to the probabilities, and these were used as inputs to the model.
More usually, treatment effects are applied to rates, rather than probabilities.
```{r combo, echo=TRUE}
# annual probabilities modified by treatment effect
pAB <- RR*nAB/nA
pAC <- RR*nAC/nC
pAD <- RR*nAD/nA
pBC <- RR*nBC/nB
pBD <- RR*nBD/nB
pCD <- RR*nCD/nC
# annual transition probability matrix
Ptc <- matrix(
  c(1.0 - pAB - pAC - pAD,               pAB,         pAC, pAD, 
                      0.0, (1.0 - pBC - pBD),         pBC, pBD,
                      0.0,               0.0, (1.0 - pCD), pCD,
                      0.0,               0.0,         0.0, 1.0),
  nrow = 4L, byrow = TRUE, 
  dimnames=list(
    source = c("A", "B", "C", "D"), target = c("A", "B", "C", "D")
  )
)
# create Markov states for combination therapy
sAc <- MarkovState$new("A", cost = dmca + ccca + cAZT + cLam)
sBc <- MarkovState$new("B", cost = dmcb + cccb + cAZT + cLam)
sCc <- MarkovState$new("C", cost = dmcc + cccc + cAZT + cLam)
sDc <- MarkovState$new("D", cost = 0.0, utility = 0.0)
# create transitions
tAAc <- Transition$new(sAc, sAc)
tABc <- Transition$new(sAc, sBc)
tACc <- Transition$new(sAc, sCc)
tADc <- Transition$new(sAc, sDc)
tBBc <- Transition$new(sBc, sBc)
tBCc <- Transition$new(sBc, sCc)
tBDc <- Transition$new(sBc, sDc)
tCCc <- Transition$new(sCc, sCc)
tCDc <- Transition$new(sCc, sDc)
tDDc <- Transition$new(sDc, sDc)
# construct the model
m.comb <- SemiMarkovModel$new(
  V = list(sAc, sBc, sCc, sDc),
  E = list(tAAc, tABc, tACc, tADc, tBBc, tBCc, tBDc, tCCc, tCDc, tDDc),
  discount.cost = cDR / 100.0,
  discount.utility = oDR / 100.0
)
# set the probabilities
m.comb$set_probabilities(Ptc)
```

The per-cycle transition matrix for the combination therapy is as follows:
```{r tmcombo, echo=FALSE}
TM <- m.comb$transition_probabilities()
pander::pander(TM, emphasize.rownames = FALSE, justify = "lcccc")
rm(TM)
```

In this model, lamivudine is given for the first 2 years, with 
the treatment effect assumed to persist for the same period. The
state populations and cycle numbers are retained by the model between 
calls to `cycle` or `cycles` and can be retrieved by calling `get_populations`.
In this example, the combination therapy model is run for 2 cycles, then the
population is used to continue with the monotherapy model for the remaining
8 years. The `reset` function is used to set the cycle number and elapsed
time of the new run of the mono model.
```{r combo_run, echo=TRUE}
# run combination therapy model for 2 years
populations <- c("A" = N, "B" = 0L, "C" = 0L, "D" = 0L)
m.comb$reset(populations)
# run 2 cycles
MT.comb <- m.comb$cycles(2L, hcc.pop = FALSE, hcc.cost = FALSE)
# feed populations into mono model & reset cycle counter and time
populations <- m.comb$get_populations()
m.mono$reset(
  populations, 
  icycle = 2L, 
  elapsed = as.difftime(365.25*2.0, units = "days")
)
# and run model for next 18 years
MT.comb <- rbind(
  MT.comb, m.mono$cycles(ncycles = 18L, hcc.pop = FALSE, hcc.cost = FALSE)
)
```

The Markov trace for combination therapy is as follows:
```{r echo=F}
keep <- c("Years", "A", "B", "C", "D", "Cost", "QALY")
pander::pander(MT.comb[, keep], row.names = FALSE, justify = "rrrrrrr", 
               round = c(2L, 0L, 0L, 0L, 0L, 0L, 3L))
```

## Comparison of treatments
```{r ICER, echo=FALSE}
el.comb <- sum(MT.comb$QALY)
cost.comb <- sum(MT.comb$Cost)
icer <- (cost.comb - cost.mono) / (el.comb - el.mono)
```
The ICER is calculated by running both models and calculating the incremental
cost per life year gained. Over the 20 year time horizon, the expected life
years gained for monotherapy was `r round(el.mono,3)` years at a total cost
per patient of `r format(cost.mono,digits=2,scientific=F,big.mark=',')` GBP.
The expected life years gained with combination therapy  for two years was
`r round(el.comb,3)` at a total cost per patient of 
`r format(cost.comb,digits=2,scientific=F,big.mark=',')` GBP. The incremental 
change in life years was `r round(el.comb-el.mono,3)` years at an incremental 
cost of `r format(cost.comb-cost.mono,digits=2,scientific=F,big.mark=',')` GBP,
giving an ICER of `r round(icer,2)` GBP/QALY. This is consistent with the
result obtained by Briggs *et al* [-@briggs2006] (6276 GBP/QALY), within 
rounding error.

# References
