---
title: "Tegaderm CHG IV Securement Dressing for Central Venous and Arterial Catheter Insertion Sites"
subtitle: "A decision tree example with probabilistic sensitivity analysis"
author: "Andrew J. Sims"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Tegaderm CHG IV Securement Dressing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: briggs:2002a
  type: book
  author:
    - family: Briggs
      given: Andrew
    - family: Claxton
      given: Karl
    - family: Sculpher
      given: Mark
  title: "Decision modelling for health economic evaluation"
  publisher: "Oxford University Press"
  publisher-place: "Oxford, UK"
  isbn: 978-0-19-852662-9
  issued:
    - year: 2006
- id: jenks:2016a
  type: article
  author:
    - given: Michelle
      family: Jenks
    - given: Joyce
      family: Craig
    - given: William
      family: Green
    - given: Neil
      family: Hewitt
    - given: Mick
      family: Arber
    - given: Andrew J.
      family: Sims
  title: "Tegaderm CHG IV Securement Dressing for Central Venous and Arterial
          Catheter Insertion Sites: A NICE Medical Technology Guidance"
  container-title: "Applied Health Economics and Health Policy"
  issued:
    - year: 2016
  volume: "14"
  page: 135-149
---
```{r, include = FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=F}
#library('rdecision')
source('rdecision-dev.R')
options(warn=0)
```

# Introduction
This vignette is an example of modelling a decision tree using the `rdecision`
package, with probabilistic sensitivity analysis. It is based on the model reported
by Jenks *et al* [-@jenks:2016a] in which a transparent dressing used to secure
vascular catheters (Tegaderm CHG) was compared with a standard dressing.

# Creating the model

## Model variables
The following code creates the variables to be used in the model. The choice of
variables, their distributions and their parameters are taken from table 4 
of Jenks *et al* [-@jenks:2016a]. 
```{r variables, echo=T}
# clinical variables
r.CRBSI <- GammaModelVariable$new(
  'Baseline CRBSI rate', 
  '/1000 catheter days',
  alpha=1.48*100, beta=1/100
)
hr.CRBSI <- LogNormalModelVariable$new(
  'Tegaderm CRBSI HR',
  'ratio',
  mu=log(0.402), sigma=0.393
)
r.LSI <- GammaModelVariable$new(
  'Baseline LSI rate',
  '/1000 catheter days',
  alpha=1.4, beta=0.1
)
hr.LSI <- LogNormalModelVariable$new(
  'Tegaderm LSI HR',
  'ratio',
  mu=log(0.402), sigma=0.393
)
r.Dermatitis <- BetaModelVariable$new(
  'Baseline dermatitis probability',
  'probability',
  alpha=1, beta=475
)
rr.Dermatitis <- LogNormalModelVariable$new(
  'Tegaderm Dermatitis RR',
  'ratio',
  mu=log(1), sigma=0.5
)

# cost variables
c.CRBSI <- GammaModelVariable$new('CRBSI cost', 'GBP', alpha=9990/3000, beta=3000)
c.Dermatitis <- GammaModelVariable$new('Dermatitis cost', 'GBP', alpha=2, beta=3)
c.LSI <- GammaModelVariable$new('LSI cost', 'GBP', alpha=100/30, beta=30)
c.Tegaderm <- ConstModelVariable$new('Tegaderm CHG cost', 'GBP', const=6.26)
c.Standard <- ConstModelVariable$new('Standard dressing cost', 'GBP', const=1.54)
n.cathdays <- GammaModelVariable$new('No. days with catheter', 'days', alpha=10/5, beta=5)
n.dressings <- GammaModelVariable$new('No. dressings', 'dressings', alpha=3/2, beta=2)
```


## Model variable expressions
Variables in the model may be included in the decision tree via model variable expressions,
which are mathematical expressions which involve model variables. The most simple
form of model variable expression involving a model variable `X` is `quote(X)`. More
complex forms of expression involving R's numerical functions and
multiple model variables are supported, provided the expressions conform to R syntax.

The following code creates the model variable expressions to be used as values in the
decision tree nodes.

```{r expressions, echo=T}
# probabilities
p.Dermatitis.S <- ModelVariableExpression$new(
  quote(r.Dermatitis),
  'Probability of dermatitis with standard dressing'
)
p.Dermatitis.T <- ModelVariableExpression$new(
  quote(r.Dermatitis*rr.Dermatitis),
  'Probability of dermatitis with Tegaderm'
)
p.LSI.S <- ModelVariableExpression$new(
 quote(r.LSI),
 'Probability of LSI with standard dressing'
)
p.LSI.T <- ModelVariableExpression$new(
 quote(r.LSI*hr.LSI),
 'Probability of LSI with Tegaderm'
)
p.CRBSI.S <- ModelVariableExpression$new(
  quote(r.CRBSI*n.cathdays/1000),
  'Probability of CRBSI with standard dressing'
)
p.CRBSI.T <- ModelVariableExpression$new(
  quote(r.CRBSI*n.cathdays*hr.CRBSI/1000),
  'Probability of CRBSI with standard dressing'
)

# costs
c.Dermatitis.MVE <- ModelVariableExpression$new(
  quote(c.Dermatitis),
  'Cost of dermatitis'
)
c.LSI.MVE <- ModelVariableExpression$new(
  quote(c.LSI),
  'Cost of LSI'
)
c.CRBSI.MVE <- ModelVariableExpression$new(
  quote(c.CRBSI),
  'Cost of CRBSI'
)
c.S <- ModelVariableExpression$new(
  quote(n.dressings*c.Standard)
)
c.T <- ModelVariableExpression$new(
  quote(n.dressings*c.Tegaderm)
)
```

## Constructing the decision tree
The following code constructs the decision tree, node by node, based on figure 2
of Jenks *et al* [-@jenks:2016a]. In the
formulation used by `rdecision`, each node is a potentially recursive
structure which is allowed to have zero or more child nodes; any child
nodes must have already been declared before their parent node is declared.
This implies that a tree should be constructed from right to left, starting
with leaf nodes which have no children (leaf nodes are synonymous with
pathways in Briggs' terminology [-@briggs:2002a]). The final node to be constructed 
is the node representing the decision problem.

```{r tree, echo=T}
# standard dressing branch
leaf.S.Dermatitis <- LeafNode$new('Dermatitis (Standard Dressing)')
leaf.S.LSI <- LeafNode$new('Local site infection (Standard Dressing)')
leaf.S.CRBSI <- LeafNode$new('CRBSI (Standard Dressing)')
leaf.S.NoComp <- LeafNode$new('No complication (Standard Dressing)')

chance.S <- ChanceNode$new(
  children = list(leaf.S.Dermatitis, leaf.S.LSI, leaf.S.CRBSI, leaf.S.NoComp),
  edgelabels = c('Dermatitis', 'Local site infection', 'CRBSI', 'No complication'),
  costs = list(c.Dermatitis.MVE, c.LSI.MVE, c.CRBSI.MVE, 0),
  p = list(p.Dermatitis.S, p.LSI.S, p.CRBSI.S, as.numeric(NA)),
  ptype = 'MVE'
)

# Tegaderm dressing branch
leaf.T.Dermatitis <- LeafNode$new('Dermatitis (Tegaderm CHG)')
leaf.T.LSI <- LeafNode$new('Local site infection (Tegaderm CHG)')
leaf.T.CRBSI <- LeafNode$new('CRBSI (Tegaderm CHG)')
leaf.T.NoComp <- LeafNode$new('No complication (Tegaderm CHG)')

chance.T <- ChanceNode$new(
  children = list(leaf.T.Dermatitis, leaf.T.LSI, leaf.T.CRBSI, leaf.T.NoComp),
  edgelabels = c('Dermatitis', 'Local site infection', 'CRBSI', 'No complication'),
  costs = list(c.Dermatitis.MVE, c.LSI.MVE, c.CRBSI.MVE, 0),
  p = list(p.Dermatitis.T, p.LSI.T, p.CRBSI.T, as.numeric(NA)), 
  ptype='MVE'
)

# decision node
d <- DecisionNode$new(
  children = list(chance.S, chance.T),
  edgelabels = c('Standard Dressing', 'Tegaderm CHG'),
  costs = list(c.S, c.T)
)

```
In the manufacturer's model, the uncertainties in the probabilities associated
with the polytomous chance nodes were modelled as independent variables. This is
not recommended because there is a chance that a particular run of the PSA will
yield probabilities that are outside the range [0,1]. Representing the 
uncertain probabilities with draws from a Dirichlet distribution is preferred. Creating
a `ChanceNode` with ModelVariableExpressions is permitted, but results
in a warning being issued.

# Documenting the model
Package `rdecision` includes tools for automated documentation of the model structure
and inputs. 

## Model inputs
The model variables associated with a node and its descendants can be tabulated
using the method `tabulateModelVariables`. The method
returns a data frame describing each variable, its description, units and uncertainty
distribution. Variables inheriting from type `ModelVariable` will be included in the
tabulation; regular numeric values will not be listed. For extensive models, variables
associated with separate branches of a tree can be tabulated separately by calling
the method for different head nodes. Selected columns are omitted from the 

```{r modelinputs, echo=T}
local({
  DF <- d$tabulateModelVariables()
  DF$Variable <- NULL
  DF$SD <- NULL
  knitr::kable(DF, row.names=F, format.args=list(scientific=F), digits=3)
})
```

# Running the model
The following code runs a single model scenario, using the `evaluate`
method of a decision node to evaluate each pathway and decision option. In 
the model there are eight possible root-to-leaf paths, each of which
begins with the decision node and ends with a leaf node. For example,
pathway `Dermatitis(Standard Dressing)` involves a traversal of nodes
`d`, `chance.S`, and `leaf.S.Dermatitis`. This model did not consider 
utility, and the columns associated with utility are removed.

```{r basecase, echo=T}
RES <- d$evaluate(expected=T)
RES$Utility <- NULL
RES$ExpectedUtility <- NULL
```

# Model results

## Base case
The results of the scenario model, using the code from the previous section,
yields the following result:
```{r basecase-tabulate, echo=F}
knitr::kable(RES)
```

There are, as expected, eight root-to-leaf pathways, which `path.apply` has 
worked out itself from the model structure. The total probability and expected
cost for each choice can be calculated from the table as follows which gives 
a result consistent with that reported by Jenks *et al* [-@jenks:2016a].

```{r basecase-summary, echo=T}
local({
  SUM <- aggregate(
    RES[,c('Probability', 'ExpectedCost')],
    by = list(RES$Choice),
    FUN = sum  
  )
  names(SUM) <- c('Choice', 'Probability', 'Expected Cost')
  knitr::kable(SUM)
})
```

## Probabilistic senstivity analysis
When they are created, each `ModelVariable` returns its expected value when its
method `value()` is called. Calling the method `sample()` of a model variable causes
it to sample from its uncertainty distribution, and return the sampled value when
method `value()` is next called. The same sampled value will be returned until
`sample()` is called again. Calling `sample(expected=T)` causes `value()` to 
return the expected value of the variable. 

Probabilistic sensitivity analysis
is supported through the use of sampling model variables. In practice, because the
model variables are contained within model variable expressions embedded in the
model, decision nodes expose a method called `sample()`. This causes each model
variable associated with the decision node and its descendants to be sampled. The code
needed to run $N$ samples of the model, and the results of the first 10 samples are 
as follows:

```{r PSA, echo=T}
local({
  N <- 10000
  RES <<- data.frame(
    Run = 1:N,
    Tegaderm = numeric(length=N),
    Standard = numeric(length=N)
  )
  for (i in 1:N) {
    RUN <- d$evaluate(expected=F)
    RES$Tegaderm[i] <<- sum(RUN$ExpectedCost[RUN$Choice=='Tegaderm CHG'])
    RES$Standard[i] <<- sum(RUN$ExpectedCost[RUN$Choice=='Standard Dressing'])
  }
  RES$Difference <<- RES$Tegaderm - RES$Standard
  knitr::kable(head(RES, n=10))
})
```

From PSA, the median cost of treatment with Tegaderm was `r round(mean(RES$Tegaderm),2)`,
the median cost of treatment with standard dressings was `r round(mean(RES$Standard),2)`
and the median cost saving was `r round(mean(RES$Difference),2)`. The 95% confidence 
interval for cost saving was `r round(quantile(RES$Difference, probs=c(0.025)),2)` to 
`r round(quantile(RES$Difference, probs=c(0.975)),2)`. 
Overall, `r round(100*sum(RES$Difference<0)/nrow(RES),2)`% of runs found that Tegaderm
was cost saving. These results replicate those reported by the manufucturer.

```{r echo=F}
rm(RES)
```

# References
